# Introduction to Convex Optimization {#sec-intro}

<!-- Conceptual on-ramp, adapted from the 2019 useR! tutorial introduction.
     Mostly prose + one self-contained convexity-chord plot (no image asset). -->

::: callout-note
## What this chapter is for
The big picture before we touch code: what an optimization problem is, what makes
one **convex**, why convexity is such good news, and where these problems show up in
statistics and machine learning.
:::

```{r}
#| label: setup
#| include: false
library(ggplot2)
```

## Optimization is everywhere in statistics

Estimation is often optimization. We **maximize** a likelihood, **minimize** a loss, or
**penalize** a fit to trade bias for variance. Ordinary least squares, ridge and
lasso, logistic regression, support vector machines, maximum-entropy models,
portfolio selection --- each is "find the parameter that makes some criterion as
small (or large) as possible, perhaps subject to constraints."

The problems we study in this book have the form

$$
\begin{array}{lll}
  \mbox{minimize}   & f_0(x) & \\
  \mbox{subject to} & f_i(x) \leq 0, & i = 1,\ldots,m\\
                    & Ax = b, &
\end{array}
$$

with variable $x \in \mathbf{R}^n$. Here $f_0$ is the **objective**, the $f_i \le 0$
are **inequality constraints**, and $Ax = b$ are **equality constraints**. (To
maximize, minimize the negative.)

## What makes it convex

A problem is **convex** when the objective and inequality functions are convex and
the equalities are linear. A function is **convex** if the line segment (chord)
joining any two points on its graph lies *on or above* the graph:

$$f\big(\theta x + (1-\theta) y\big) \;\le\; \theta f(x) + (1-\theta) f(y),
\qquad \theta \in [0,1].$$

```{r}
#| label: chord
#| echo: false
#| fig-width: 5
#| fig-height: 3.2
#| fig-cap: "A convex function curves upward: every chord lies above the graph."
f <- function(x) 0.5 * x^2
xs <- seq(-2.2, 2.2, length.out = 200)
a <- -1.6; b <- 1.3
ggplot(data.frame(x = xs, y = f(xs)), aes(x, y)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_segment(aes(x = a, y = f(a), xend = b, yend = f(b)),
               color = "red", linewidth = 0.8) +
  geom_point(data = data.frame(x = c(a, b), y = f(c(a, b))), color = "red") +
  theme_minimal() + labs(x = NULL, y = NULL)
```

Picture it as a bowl. A **concave** function is an upside-down bowl ($-f$ is convex);
an **affine** (flat) function is both. We minimize convex functions and maximize
concave ones.

## Why convexity is such good news

Convex problems have properties that make them a joy to work with:

- **Every local optimum is global.** No getting stuck in a bad valley --- if you find
  a solution, it is *the* solution.
- **The feasible region is convex** (an intersection of convex sets), so the search
  space is well-behaved.
- They are **solved reliably and fast** by mature algorithms, even when the problem is
  nonlinear or nondifferentiable, and at scale.

This is exactly why it is worth recognizing --- or reformulating into --- a convex
problem: you trade the uncertainty of general optimization for a global guarantee.

## Convex problems in statistics and machine learning

A small sample of methods that are, under the hood, convex optimization:

- ordinary, nonnegative, and constrained **least squares**;
- **ridge, lasso, elastic-net**, and other penalized regressions;
- **logistic regression** and **support vector machines**;
- **quantile** and **Huber (robust)** regression;
- **isotonic** and **shape-constrained** fitting;
- **maximum-entropy** and **log-concave density** estimation;
- **portfolio** optimization and many problems in finance, signal processing, and control.

New methods join the list every year --- and CVXR lets you build them without writing
a custom solver.

## What about non-convex problems?

Not everything is convex --- in fact most problems are not, and they are harder in
general (no global guarantee). Two responses recur in this book:

- A non-convex problem often has a **convex relaxation** --- an approximation you *can*
  solve that yields a useful bound or solution.
- Some problems are non-convex but **smooth**, and a nonlinear solver will find a good
  local solution; CVXR handles these too (@sec-dnlp).

## The plan

The rest of the book is hands-on. We start gently --- your first CVXR problems
(@sec-gentle-intro) --- then learn the one rule that governs what CVXR will accept
(@sec-dcp), survey the landscape of problem types (@sec-problem-types), and work
through regression, classification, finance, integer, and nonlinear examples. By the
end you will translate a problem from a paper or textbook directly into working R.

Read on --- and run the code as you go.
