# A Gentle Introduction {#sec-gentle-intro}

<!-- Source of truth: cvxr_docs/examples/gentle-intro.qmd + 2019 ch.05. -->

::: callout-note
## What you'll be able to do
By the end of this chapter you can state a convex problem in CVXR, solve it, read
its status and solution, modify and re-solve it, and recognize when a problem is
*not* something CVXR will accept.
:::

```{r}
#| label: setup
#| include: false
library(CVXR)
source("_check_status.R")
```

## Convex optimization in one picture

A **convex optimization problem** has the form

$$
\begin{array}{ll}
  \mbox{minimize}   & f_0(x)\\
  \mbox{subject to} & f_i(x) \leq 0, \quad i = 1,\ldots,m\\
                    & g_i(x) = 0,    \quad i = 1,\ldots,p
\end{array}
$$

where $x$ is the variable, the objective $f_0$ and the inequality functions
$f_1,\ldots,f_m$ are **convex**, and the equality functions $g_1,\ldots,g_p$ are
**affine**.

A handy mental picture:

- a **convex** function is *upward-curving* (a chord lies above the graph),
- a **concave** function is *downward-curving*,
- an **affine** function is *flat* --- both convex and concave.

We minimize a convex function (or maximize a concave one) over a convex region.
The payoff for staying convex: **any local solution is automatically global.**
For much more, see @BoydVandenberghe:2004.

## 'Hello World'

Here is about the simplest problem that shows all three ingredients --- an
objective, an inequality constraint, and an equality constraint:

$$
\begin{array}{ll}
  \mbox{minimize}   & x^2 + y^2 \\
  \mbox{subject to} & x \geq 0, \quad 2x + y = 1
\end{array}
$$

with scalar variables $x$ and $y$. It is simple enough to solve by hand, so we can
check CVXR's answer. Here it is in CVXR:

```{r}
#| label: hello
# Variables minimized over
x <- Variable(1)
y <- Variable(1)

# Problem definition
objective   <- Minimize(x^2 + y^2)
constraints <- list(x >= 0, 2 * x + y == 1)
prob1 <- Problem(objective, constraints)

# Solve it
sol1 <- psolve(prob1)

status(prob1)   # did it solve?
sol1            # optimal objective value
value(x)
value(y)
```

Let's walk through that carefully, because every CVXR problem is built the same way.

**Variables are placeholders.** The first two lines create scalar `Variable`
objects. They have no value yet --- not until we solve.

```{r}
#| eval: false
x <- Variable(1)
y <- Variable(1)
```

**`Minimize()` builds an objective, not a number.** Unlike base R's `min()`, this
does not compute anything (there is nothing to compute --- `x` and `y` have no
values). It records the *goal*.

```{r}
#| eval: false
objective <- Minimize(x^2 + y^2)
```

**`>=` and `==` build constraints, not `TRUE`/`FALSE`.** The comparison operators
are overloaded: `x >= 0` returns a *Constraint* object the solver will enforce ---
not a logical. (Without the constraints, the solution would trivially be
$x = y = 0$.)

```{r}
#| eval: false
constraints <- list(x >= 0, 2 * x + y == 1)
```

**`Problem()` assembles, `psolve()` solves.** A problem pairs an objective with
zero or more constraints. `psolve()` translates it into a form a convex solver
understands, calls the solver, and returns the optimal objective value. Afterwards
`value()` retrieves each variable and `status()` reports what happened.

Solving by hand: the equality gives $y = 1 - 2x$, so we minimize
$x^2 + (1-2x)^2 = 5x^2 - 4x + 1$, which is smallest at $x = 2/5$. Then $y = 1/5$ and
the objective is $1/5$ --- exactly what CVXR found. The world says 'hi' back.

### Three things that can happen

When you `psolve()` a problem, `status()` is one of:

1. **`"optimal"`** --- solved; `value()` gives the optimal variables.
2. **`"infeasible"`** --- no point satisfies all constraints (e.g. `x >= 1` *and*
   `x <= 0`). `psolve()` returns `+Inf` (minimization) and variable values are `NA`.
3. **`"unbounded"`** --- the objective can be driven to $-\infty$. `psolve()`
   returns `-Inf`.

Always glance at `status()` before trusting a solution.

## Watching the solver work

By default `psolve()` is quiet. But you can ask it to **show its work** by passing
`verbose = TRUE` --- invaluable when you want to understand what CVXR is doing, or
diagnose a slow or stalling solve:

```{r}
#| label: verbose
#| message: true
psolve(prob1, verbose = TRUE)
```

The output has three parts, and they preview the rest of this book:

1. **Compilation** --- CVXR translates your problem into a standard form a solver
   understands. The chain (here `Dcp2Cone -> ... -> ConeMatrixStuffing -> ...`) is
   the machinery we open up in @sec-dcp; for now, note that *your symbolic problem
   becomes solver data*.
2. **Numerical solver** --- the chosen solver's own iteration log, converging to the
   answer.
3. **Summary** --- the status, optimal value, and how long compilation vs. solving
   took.

Turn it on whenever you are curious or stuck; turn it off (the default) once things
work. Every `psolve()` call in this book accepts it.

## Modifying and re-solving

CVXR objects are ordinary R objects: you can build new problems from old ones. Here
we reuse the constraints from `prob1`, change the objective, and add a constraint:

```{r}
#| label: modify
prob2 <- Problem(Minimize(x^2 + y^2 + abs(x - y)),
                 c(prob1@constraints, list(y <= 1)))
sol2 <- psolve(prob2)
status(prob2)
sol2
value(x)
value(y)
```

This *modify-and-re-solve* loop --- change one thing, solve again --- is what makes
CVXR so good for prototyping new methods.

## An *invalid* problem

You can't type *any* problem you like. Suppose we tried to **maximize** the convex
objective $x^2 + y^2$:

```{r}
#| label: invalid
#| error: true
prob_bad <- Problem(Maximize(x^2 + y^2), prob1@constraints)
psolve(prob_bad)
```

CVXR refuses. Maximizing a convex function is *not* a convex problem, and CVXR
checks this using a strict rule set called **Disciplined Convex Programming (DCP)**.
Follow the rules and your problem is guaranteed convex; break them and CVXR stops
you rather than silently returning nonsense. The next chapter is devoted to DCP ---
this error is your first encounter with it.

## A familiar example: least squares

Let's connect to something every statistician knows. We generate data from a known
linear model $Y = X\beta + \epsilon$ and recover $\beta$.

```{r}
#| label: ols-data
set.seed(123)
n <- 50; p <- 10
beta_true <- -4:5                      # the true coefficients, -4 through 5
X <- matrix(rnorm(n * p), nrow = n)
Y <- X %*% beta_true + rnorm(n)
```

The textbook tool is `lm()`:

```{r}
#| label: ols-lm
ls_fit <- lm(Y ~ 0 + X)                # no intercept in our model
coef(ls_fit)
```

In CVXR we write the least-squares problem in its mathematical form,
$\min_\beta \|Y - X\beta\|_2^2$:

```{r}
#| label: ols-cvxr
beta <- Variable(p)
obj  <- Minimize(sum((Y - X %*% beta)^2))
ols  <- Problem(obj)
sol  <- psolve(ols)
```

Notice the objective is written with the same `sum`, `%*%`, and `^` you use on
ordinary R numbers --- but `beta` is a CVXR variable. The estimates match `lm()`:

```{r}
#| label: ols-compare
cbind(CVXR = value(beta), lm = coef(ls_fit))
```

### Anything is a function of the solution

A solved problem gives you more than the coefficients. Once `beta` has a value, you
can evaluate **any expression built from it** with `value()`. The **fitted values**
$X\hat\beta$ and the residuals, for instance, are one line each:

```{r}
#| label: ols-fitted
fitted <- value(X %*% beta)        # X times beta-hat
resid  <- Y - fitted
head(round(cbind(fitted, resid), 3))
```

`X %*% beta` is a CVXR *expression*; `value()` evaluates it at the solution, and the
result is ordinary R you can plot, summarize, or feed onward. This little move ---
*any quantity you care about is `value()` of an expression* --- is one we will reuse in
every chapter.

### So what did we gain?

Fair question --- we replaced one call to `lm()` with several lines, and it even
runs slower. The payoff appears the moment the problem changes. Suppose we *know*
the coefficients are **nonnegative**. That is
[nonnegative least squares](https://en.wikipedia.org/wiki/Non-negative_least_squares),
and `lm()` can no longer help. In CVXR it is **one extra constraint**:

```{r}
#| label: nnls
nnls_prob <- Problem(obj, constraints = list(beta >= 0))
psolve(nnls_prob)
value(beta)
```

We can check against the dedicated `nnls` package:

```{r}
#| label: nnls-check
cbind(CVXR = value(beta), nnls = nnls::nnls(X, Y)$x)
```

Same answer --- but in CVXR we got there by *adding a line*, not by finding a new
package. That is the whole idea: you describe the problem, CVXR solves it.

## Exercises

Try these by editing the chunks above and re-running. Solutions are one click away.

**Exercise 1.** Change the least-squares objective to **least absolute deviation**
(minimize the sum of absolute residuals) and solve. Which objective gives a smaller
*sum of absolute* residuals?

::: {.callout-tip collapse="true"}
## Solution
```{r}
#| label: ex-lad
lad_obj  <- Minimize(sum(abs(Y - X %*% beta)))
lad_prob <- Problem(lad_obj)
psolve(lad_prob)
cbind(LAD = value(beta), lm = coef(ls_fit))
```
LAD minimizes absolute residuals, so by that yardstick it wins; `lm` minimizes
*squared* residuals and would win on sum of squares. Note we reused `beta` --- its
value reflects whichever problem was last solved.
:::

**Exercise 2.** Add the constraint that the coefficients are **nondecreasing**:
$\beta_1 \le \beta_2 \le \cdots \le \beta_p$.

*Hint:* what base-R function takes lagged differences?

::: {.callout-tip collapse="true"}
## Solution
`base::diff()` has a CVXR method, so the monotonicity constraint is just
`diff(beta) >= 0`:
```{r}
#| label: ex-monotone
mono <- Problem(obj, constraints = list(diff(beta) >= 0))
psolve(mono)
value(beta)
```
:::

**Exercise 3 (⭐).** Constrain the first four coefficients to sum to at least 4,
$\sum_{i=1}^{4}\beta_i \ge 4$, using a matrix row rather than writing out four terms.

::: {.callout-tip collapse="true"}
## Solution
Build a row vector that picks out $\beta_1,\ldots,\beta_4$:
```{r}
#| label: ex-sum
A <- matrix(c(rep(1, 4), rep(0, 6)), nrow = 1)
sum_con <- Problem(obj, constraints = list(A %*% beta >= 4))
psolve(sum_con)
value(beta)
```
Working with matrices generalizes far better than writing `beta[1] + beta[2] + ...`.
:::

## Takeaways

- Every CVXR problem is **Variable → objective → constraints → `Problem` →
  `psolve`**, then `status()` and `value()`.
- Operators like `>=`, `==`, `sum`, `%*%` are *overloaded* to build optimization
  objects, not to compute numbers.
- The power is **flexibility**: change the loss, add a constraint, re-solve --- no
  closed form or special-purpose package required.
- CVXR only accepts problems that pass the **DCP** rules, which we turn to next.

For a longer version of this introduction, see the
[CVXR website](https://cvxr.rbind.io/examples/gentle-intro.html).
