# Penalized Regression: Ridge, Lasso, Elastic Net {#sec-elastic-net}

<!-- LIVE. Source: cvxr_docs/examples/regression/elastic-net.qmd (+ ridge/lasso,
     glmnet-equivalence). First rung of the regression spine; foreshadows Ch.4
     (change the loss) and Ch.5 (DPP sweep). -->

::: callout-note
## What you'll be able to do
Add an $\ell_2$ (ridge), $\ell_1$ (lasso), or combined (elastic-net) penalty to a
regression; see ridge and lasso as two ends of **one** penalty family; trace a
regularization path; and validate against `glmnet`.
:::

```{r}
#| label: setup
#| include: false
library(CVXR)
library(ggplot2)
library(tidyr)
source("_check_status.R")
```

## Why regularize?

In @sec-gentle-intro we recovered coefficients by least squares. When predictors
are many or correlated, plain least squares **overfits** --- the estimates become
large and unstable. The cure is to add a **penalty** that discourages large
coefficients. The two classic penalties are the squared $\ell_2$-norm and the
$\ell_1$-norm, and the **elastic net** combines them:

$$
\underset{\beta}{\mbox{minimize}}\quad
\frac{1}{2m}\|y - X\beta\|_2^2
\;+\;\lambda\Big(\tfrac{1-\alpha}{2}\|\beta\|_2^2 + \alpha\|\beta\|_1\Big).
$$

Here $\lambda \ge 0$ sets the overall strength and $\alpha\in[0,1]$ dials between
the two penalties. The family contains both classics:

- $\alpha = 0$ → **ridge** (shrinks coefficients smoothly),
- $\alpha = 1$ → **lasso** (drives some coefficients exactly to zero),
- $0 < \alpha < 1$ → **elastic net** (does both; handles correlated groups).

## One modular penalty

Because CVXR lets us write the penalty as an ordinary R function, the *whole family*
is a few lines:

```{r}
#| label: reg-fn
elastic_reg <- function(beta, lambda = 0, alpha = 0) {
  ridge <- (1 - alpha) / 2 * sum_squares(beta)
  lasso <- alpha * p_norm(beta, 1)
  lambda * (lasso + ridge)
}
```

We will add it to a least-squares loss. (Keep an eye on this `loss` line --- in
@sec-change-the-loss we change *only that line* to get robust and quantile
regression for free.)

## Some correlated, sparse data

Correlated features are the classic reason to prefer elastic net over pure lasso:
lasso picks one variable from a correlated group arbitrarily, while elastic net
lets the group enter or leave together. We build a Toeplitz correlation
($\rho = 0.8$), $n = 100$, $p = 50$, with 10 truly non-zero coefficients.

```{r}
#| label: data
set.seed(1)
p <- 50; n <- 100
rho <- 0.8
Sigma <- rho^abs(outer(seq_len(p), seq_len(p), "-"))
X <- MASS::mvrnorm(n, mu = rep(0, p), Sigma = Sigma)

beta_true <- rep(0, p)
beta_true[c(1, 5, 10, 15, 20, 25, 30, 35, 40, 45)] <-
  c(3, -2.5, 2, -1.5, 1, -3, 2.5, -2, 1.5, -1)

signal_sd <- sqrt(drop(t(beta_true) %*% Sigma %*% beta_true))
y <- X %*% beta_true + rnorm(n, sd = signal_sd / 3)
y <- y / sqrt(mean(y^2))   # scale so sqrt(mean(y^2)) = 1 (see glmnet note)
```

## Trace a regularization path

We fit the elastic net ($\alpha = 0.75$) across a grid of $\lambda$, collecting the
coefficients at each. This is the **modify-and-re-solve** loop again, in a `sapply`:

```{r}
#| label: path
lambda_vals <- 10^seq(-2, 0, length.out = 10)
beta <- Variable(p)
loss <- sum_squares(y - X %*% beta) / (2 * n)
alpha <- 0.75

beta_vals <- sapply(lambda_vals, function(lambda) {
  prob <- Problem(Minimize(loss + elastic_reg(beta, lambda, alpha)))
  psolve(prob)
  check_solver_status(prob)
  value(beta)
})
```

As $\lambda$ grows, coefficients shrink toward zero --- the **regularization path**:

```{r}
#| label: path-plot
#| fig-cap: "Elastic-net regularization path."
df_path <- data.frame(lambda = lambda_vals, t(beta_vals)) |>
  pivot_longer(-lambda, names_to = "Coefficient", values_to = "Value")
ggplot(df_path, aes(lambda, Value, color = Coefficient)) +
  geom_line() + scale_x_log10() +
  labs(x = expression(lambda), y = expression(beta)) +
  theme_minimal() + guides(color = "none")
```

## Validate against `glmnet`

A reassuring check: with $y$ scaled so its root-mean-square is 1 (a `glmnet`
standardization subtlety), the same $\lambda$ grid gives `glmnet` and CVXR the same
fit. We compare at the strongest penalty:

```{r}
#| label: glmnet
library(glmnet)
g <- glmnet(X, y, family = "gaussian", alpha = alpha,
            lambda = lambda_vals, standardize = FALSE, intercept = FALSE)
nz <- which(beta_true != 0)
cbind(CVXR   = round(beta_vals[nz, 10], 3),
      glmnet = round(as.numeric(coef(g)[-1, 1])[nz], 3))
```

Close agreement --- and CVXR got there from the penalty written in plain math,
with no special-purpose path algorithm.

## Exercise

**Exercise.** Fit pure **ridge** ($\alpha = 0$) and pure **lasso** ($\alpha = 1$)
at $\lambda = 0.3$. How many coefficients are *exactly* zero under each? (This is
the defining difference between them.)

::: {.callout-tip collapse="true"}
## Solution
```{r}
#| label: ex-ridge-lasso
fit_at <- function(a, lambda = 0.3) {
  prob <- Problem(Minimize(loss + elastic_reg(beta, lambda, a)))
  psolve(prob); value(beta)
}
ridge_b <- fit_at(0)
lasso_b <- fit_at(1)
c(ridge_zeros = sum(abs(ridge_b) < 1e-6),
  lasso_zeros = sum(abs(lasso_b) < 1e-6))
```
Ridge shrinks everything but zeroes (almost) nothing; lasso sets many coefficients
exactly to zero --- that is what makes it a *variable selector*.
:::

## Takeaways

- Ridge, lasso, and elastic net are **one family**, selected by $\alpha$ --- and in
  CVXR, one small function.
- A regularization path is just the **same problem re-solved** over a $\lambda$ grid.
- CVXR matches `glmnet` while staying fully general: the next chapter changes the
  *loss*, and @sec-dpp makes this very sweep dramatically faster.

For more, see [Elastic Net](https://cvxr.rbind.io/examples/regression/elastic-net.html).
