# Types of Convex Optimization Problems {#sec-problem-types}

<!-- Survey/taxonomy: LP < QP < SOCP < SDP, one tiny verified example each.
     Source: 2019 ch02 + cvxr_docs basic/*. Verified vs CVXR 1.9.1. -->

::: callout-note
## What you'll be able to do
Recognize the standard families of convex problems --- **LP, QP, SOCP, SDP** --- see a
tiny CVXR example of each, and understand that you rarely have to choose: DCP
(@sec-dcp) classifies your problem and CVXR routes it to a capable solver.
:::

```{r}
#| label: setup
#| include: false
library(CVXR)
```

## A hierarchy of problem classes

Convex problems come in named families of increasing generality. Each contains the
previous one as a special case:

$$\text{LP} \;\subset\; \text{QP} \;\subset\; \text{SOCP} \;\subset\; \text{SDP}.$$

You usually do **not** pick the family by hand. You write the problem from atoms; DCP
certifies it is convex; and CVXR canonicalizes it into a cone program and hands it to a
solver that can handle the cones involved (you saw this chain in the `verbose` output
of @sec-gentle-intro). Still, it helps to recognize the families --- they tell you what
is expressible and which solvers apply.

## Linear program (LP)

Linear objective, linear constraints. The workhorse of operations research.

```{r}
#| label: lp
x <- Variable(2)
lp <- Problem(Minimize(sum(c(2, 3) * x)),
              list(x >= 0, x[1] + x[2] >= 1, x <= 5))
psolve(lp)
value(x)
```

## Quadratic program (QP)

A convex **quadratic** objective with linear constraints. Least squares (and ridge,
and the Markowitz portfolio of @sec-portfolio) live here.

```{r}
#| label: qp
set.seed(1)
A <- matrix(rnorm(20), 10, 2); b <- rnorm(10)
z <- Variable(2)
# constrained least squares: quadratic objective, linear constraint -> a QP
psolve(Problem(Minimize(sum_squares(A %*% z - b)), list(z >= 0)))
value(z)
```

## Second-order cone program (SOCP)

Adds constraints involving the Euclidean norm --- "second-order cones." Anything with
a `p_norm(., 2)`, `huber`, or robust-constraint flavor is typically an SOCP. Minimizing
a norm subject to a budget is the canonical toy:

```{r}
#| label: socp
xs <- Variable(3)
psolve(Problem(Minimize(p_norm(xs, 2)), list(sum(xs) == 1)))
round(value(xs), 4)             # 1/3 each -> the min-norm point on the simplex plane
```

## Semidefinite program (SDP)

The most general of the four: constraints that a **matrix is positive semidefinite**
($\succeq 0$), written with `%>>%`. Eigenvalue optimization, covariance/correlation
problems, and relaxations of hard combinatorial problems are SDPs. Here we compute the
largest eigenvalue of a matrix $M$ as $\min\{t : tI - M \succeq 0\}$:

```{r}
#| label: sdp
M <- matrix(c(2, 1, 1, 2), 2, 2)
t <- Variable(1)
psolve(Problem(Minimize(t), list((t * diag(2) - M) %>>% 0)))   # = lambda_max(M) = 3
```

Variables can themselves be matrices constrained to be PSD --- e.g. finding the nearest
unit-diagonal PSD matrix to $M$:

```{r}
#| label: sdp-var
X <- Variable(c(2, 2), PSD = TRUE)
psolve(Problem(Minimize(sum_squares(X - M)), list(diag(X) == 1)))
round(value(X), 3)
```

## And their integer cousins

Allow some variables to be **integer** or **boolean** and each family gains a
mixed-integer version (MILP, MIQP, MI-SOCP). Those are no longer convex --- they are
combinatorial --- but CVXR expresses them the same way and hands them to an
integer-capable solver. We devote @sec-mip to them.

## Takeaways

- Convex problems form a hierarchy **LP $\subset$ QP $\subset$ SOCP $\subset$ SDP**;
  each tiny example above is a few lines of CVXR.
- You write problems from **atoms**, not by naming a family --- DCP classifies and CVXR
  routes to a solver automatically.
- Adding integrality gives the mixed-integer versions (@sec-mip).

For more worked basics, see the
[CVXR examples gallery](https://cvxr.rbind.io) (least squares, LP, QP, SOCP, SDP).
