# Parameters and Efficient Sweeps (DPP) {#sec-dpp}

<!-- LIVE (CORE). Source: cvxr_docs/examples/dpp/tutorial.qmd. The climax of the
     regression spine: turn Ch.3's lambda sweep into a Parameter and reap the
     compilation-caching speedup. -->

::: callout-note
## What you'll be able to do
Replace a numeric constant with a `Parameter`, recognize when a parametrized
problem follows the **DPP** rules, and re-solve it over a grid of values far faster
than rebuilding the problem each time.
:::

```{r}
#| label: setup
#| include: false
library(CVXR)
source("_check_status.R")
```

## A hidden cost in Chapter 3

In @sec-elastic-net we traced a regularization path by building a **new problem**
for each $\lambda$ and solving it. That works, but it hides a cost: every solve
re-runs CVXR's *compilation* --- the translation of your symbolic problem into the
matrices a solver wants. When you sweep dozens or hundreds of values, you pay that
toll over and over.

**Parameters** fix this. A `Parameter` is a symbolic placeholder for a constant
whose value you can change *without rebuilding the problem*. If the problem obeys
the **Disciplined Parametrized Programming (DPP)** rules, CVXR compiles it **once**,
caches the parameter-to-data mapping, and later solves just swap in the new number.

## Parameters and the DPP rule

A `Parameter` looks like a variable but stands for data you will supply:

```{r}
#| label: param
gamma <- Parameter(nonneg = TRUE)   # a nonnegative constant, value set later
x <- Variable(5)
prob <- Problem(Minimize(sum_squares(x - 1) + gamma * p_norm(x, 1)))
is_dpp(prob)
```

DPP adds one main restriction to DCP: **you may not multiply two parametrized
expressions** (a parameter times a parameter, or a parameter times a
parameter-containing expression). Such a problem is still DCP, just not DPP --- so
it solves, but without the speedup:

```{r}
#| label: nondpp
#| results: hold
g <- Parameter(nonneg = TRUE); z <- Variable()
bad <- Problem(Minimize(g * g * z), list(z >= 1))
cat("Is DPP?", is_dpp(bad), "\n")
cat("Is DCP?", is_dcp(bad), "\n")
```

Just like the DCP reformulation skill from @sec-dcp, a non-DPP problem can usually
be **rewritten** to comply --- here by introducing a helper variable:

```{r}
#| label: dpp-fix
#| results: hold
z <- Variable(); w <- Variable(); g <- Parameter(nonneg = TRUE)
good <- Problem(Minimize(g * w), list(w == g * z, z >= 1))
cat("Is DPP?", is_dpp(good), "\n")
```

## The payoff: re-solve vs. rebuild

How much does skipping recompilation actually save? Let's measure it properly with
`bench::mark`, which reports the garbage-collection-free **minimum** time --- the
honest way to compare. We set up a lasso whose penalty weight is a `Parameter`, and
compare one **DPP re-solve** (change the value, solve again) against one **rebuild**
(fresh problem with a numeric weight).

```{r}
#| label: bench-setup
library(bench)
set.seed(1)
n <- 120; m <- 40
A <- matrix(rnorm(n * m), n, m); b <- rnorm(n)
x <- Variable(m)
error <- sum_squares(A %*% x - b)

gamma <- Parameter(nonneg = TRUE)
param_prob <- Problem(Minimize(error + gamma * p_norm(x, 1)))   # DPP

value(gamma) <- 0.5; psolve(param_prob)   # one-time compile, up front
```

```{r}
#| label: bench-run
bench::mark(
  `DPP re-solve` = { value(gamma) <- 0.7; psolve(param_prob) },
  `rebuild`      = { psolve(Problem(Minimize(error + 0.7 * p_norm(x, 1)))) },
  check = FALSE, min_iterations = 5, filter_gc = FALSE
)[, c("expression", "min", "median", "mem_alloc")]
```

On our machine the DPP re-solve is roughly **2–3× faster per solve** and allocates
less memory --- it reuses the cached compilation instead of redoing it. Over a sweep
of $K$ values you save that difference $K$ times.

::: callout-note
## How big is the win, really?
The saving is exactly the **recompilation you avoid**, so it is largest when
compilation is a meaningful share of each solve --- small-to-medium problems solved
*many* times (regularization paths, grid searches, simulations, bootstraps). For
very large problems the solver itself dominates and the ratio shrinks toward 1×.
Either way the parametrized code is cleaner, and --- as we note below --- parameters
are also the gateway to *differentiating through* a problem.
:::

## Back to the regularization path

This is exactly Chapter 3's sweep, done right. We make $\lambda$ a `Parameter` and
re-solve --- the path looks identical, but only one compilation happens:

```{r}
#| label: dpp-path
lambda_vals <- 10^seq(-3, 1, length.out = 30)
lambda <- Parameter(nonneg = TRUE)
beta <- Variable(m)
path_prob <- Problem(Minimize(error + lambda * p_norm(beta, 1)))

betas <- sapply(lambda_vals, function(l) { value(lambda) <- l; psolve(path_prob); value(beta) })
dim(betas)   # m coefficients x number of lambda values
```

Whenever you plan to solve the *same shape* of problem many times --- a path, a
grid search, a simulation, a bootstrap --- reach for a `Parameter`.

## Exercise

**Exercise (⭐).** The expression `gamma * p_norm(x, 1)` is DPP, but
`(gamma * gamma) * p_norm(x, 1)` is not. Suppose you really want the penalty weight
to be $\gamma^2$. How can you keep the problem DPP *and* sweep over $\gamma$?

::: {.callout-tip collapse="true"}
## Solution
Do the nonlinear part **in R**, outside the DSL: keep a single parameter that holds
the value $\gamma^2$, and set it from your $\gamma$ grid.
```{r}
#| label: ex-dpp
w <- Parameter(nonneg = TRUE)            # will hold gamma^2
prob <- Problem(Minimize(error + w * p_norm(x, 1)))
is_dpp(prob)
for (g in lambda_vals) { value(w) <- g^2; psolve(prob) }   # sweep, still DPP
```
The rule of thumb: parameters may enter *affinely*; any nonlinear transform of a
parameter should be computed in R and stored in another parameter.
:::

## Takeaways

- A **`Parameter`** is a constant you can change without rebuilding the problem.
- If the problem is **DPP** (`is_dpp()` --- essentially "no parameter × parameter"),
  CVXR compiles once and re-solves cheaply.
- Use it for any **repeated solve** --- paths, grid searches, simulations. It is the
  efficient way to do the sweep you met in @sec-elastic-net.

For more, see the [DPP Tutorial](https://cvxr.rbind.io/examples/dpp/tutorial.html).
