# Penalized Regression: Ridge, Lasso, Elastic Net {#sec-elastic-net}

<!-- LIVE. Structured sparse signal (10 of 50 nonzero, correlated features),
     UNSCALED y so real magnitudes are recovered. Lasso is the hero: recovers
     the support + signs + magnitudes and matches glmnet exactly (alpha=1,
     standardize=FALSE). Family (ridge/elastic) via elastic_reg() + exercise.
     Verified vs CVXR 1.9.1. -->

::: callout-note
## What you'll be able to do
Add an $\ell_2$ (ridge), $\ell_1$ (lasso), or combined (elastic-net) penalty to a
regression; watch the **lasso recover a sparse signal** and select the right
variables; validate against `glmnet`; and trace a regularization path.
:::

```{r}
#| label: setup
#| include: false
library(CVXR)
library(ggplot2)
library(tidyr)
source("_check_status.R")
```

## Why regularize?

In @sec-gentle-intro we recovered coefficients by least squares. When predictors are
many or correlated, plain least squares **overfits** --- the estimates blow up and
become unstable. The cure is a **penalty** that discourages large coefficients. The
two classics are the squared $\ell_2$-norm and the $\ell_1$-norm; the **elastic net**
combines them:

$$
\underset{\beta}{\mbox{minimize}}\quad
\frac{1}{2m}\|y - X\beta\|_2^2
\;+\;\lambda\Big(\tfrac{1-\alpha}{2}\|\beta\|_2^2 + \alpha\|\beta\|_1\Big).
$$

Here $\lambda \ge 0$ sets the strength and $\alpha\in[0,1]$ dials between the two:
$\alpha = 0$ is **ridge** (smooth shrinkage), $\alpha = 1$ is **lasso** (sets
coefficients exactly to zero --- variable selection), and in between is the **elastic
net**.

## One modular penalty

Because the penalty is just an R function, the whole family is a few lines:

```{r}
#| label: reg-fn
elastic_reg <- function(beta, lambda = 0, alpha = 0) {
  ridge <- (1 - alpha) / 2 * sum_squares(beta)
  lasso <- alpha * p_norm(beta, 1)
  lambda * (lasso + ridge)
}
```

## A structured, sparse signal

To see regularization *work*, we build data with a **known** sparse truth: only 10 of
$p = 50$ coefficients are nonzero, the features are correlated (Toeplitz, $\rho = 0.8$),
and $n = 100$ is barely above $p$ --- exactly where least squares falls apart.

```{r}
#| label: data
set.seed(1)
p <- 50; n <- 100
Sigma <- 0.8^abs(outer(seq_len(p), seq_len(p), "-"))
X <- MASS::mvrnorm(n, mu = rep(0, p), Sigma = Sigma)

beta_true <- rep(0, p)
nz <- c(1, 5, 10, 15, 20, 25, 30, 35, 40, 45)         # the true nonzero positions
beta_true[nz] <- c(3, -2.5, 2, -1.5, 1, -3, 2.5, -2, 1.5, -1)

signal_sd <- sqrt(drop(t(beta_true) %*% Sigma %*% beta_true))
y <- X %*% beta_true + rnorm(n, sd = signal_sd / 3)   # signal-to-noise ratio ~ 3
```

## Recovering the signal with the lasso

The lasso ($\alpha = 1$) both shrinks **and selects** --- it should drive the 40
irrelevant coefficients to zero and recover the 10 real ones. Fit it at a moderate
penalty:

```{r}
#| label: lasso-fit
beta <- Variable(p)
loss <- sum_squares(y - X %*% beta) / (2 * n)
lambda <- 0.15
prob <- Problem(Minimize(loss + elastic_reg(beta, lambda, alpha = 1)))
psolve(prob)
check_solver_status(prob)
```

How well did it do? Line up the estimate against the **truth** at the 10 true-nonzero
positions (plus a few that should be zero) --- and against `glmnet` while we are at it:

```{r}
#| label: recovery
library(glmnet)
g <- glmnet(X, y, family = "gaussian", alpha = 1, lambda = lambda,
            standardize = FALSE, intercept = FALSE)
data.frame(truth  = beta_true,
           CVXR   = round(value(beta), 3),
           glmnet = round(as.numeric(coef(g)[-1]), 3))[c(nz, 2, 3, 4), ]
```

The lasso recovers every true coefficient's **sign and rough magnitude**, leaves the
irrelevant ones at (or near) zero, and agrees with `glmnet` to three decimals --- a
genuine, non-trivial fit, not a vector of zeros. (Correlated features buy a few small
false positives; that is the cost of the setting.)

## The regularization path

Sweeping $\lambda$ traces each coefficient from its near-least-squares value down to
zero. As in @sec-gentle-intro this is the **same problem re-solved** over a grid ---
the place where the `Parameter` trick of @sec-dpp pays off:

```{r}
#| label: path
lambda_vals <- 10^seq(-2, 0.7, length.out = 25)
beta_vals <- sapply(lambda_vals, function(l) {
  psolve(Problem(Minimize(loss + elastic_reg(beta, l, alpha = 1))))
  value(beta)
})
```

```{r}
#| label: path-plot
#| fig-cap: "Lasso regularization path. True-nonzero coefficients (blue) persist; the rest (grey) are driven to zero."
df <- data.frame(lambda = lambda_vals, t(beta_vals)) |>
  pivot_longer(-lambda, names_to = "coef", values_to = "value")
df$true_nz <- as.integer(sub("X", "", df$coef)) %in% nz
ggplot(df, aes(lambda, value, group = coef, color = true_nz)) +
  geom_line(alpha = 0.8) +
  geom_vline(xintercept = lambda, linetype = "dashed") +
  scale_x_log10() +
  scale_color_manual(values = c("grey75", "steelblue"),
                     labels = c("true zero", "true nonzero"), name = NULL) +
  labs(x = expression(lambda), y = expression(hat(beta))) +
  theme_minimal()
```

The blue (truly nonzero) coefficients stay large longest; the grey ones collapse to
zero quickly. The dashed line marks the $\lambda$ we fit above.

## Exercise

**Exercise.** Refit the same data with **ridge** ($\alpha = 0$) at `lambda = 0.15` and
compare with the lasso fit. How many coefficients does each set exactly to zero --- and
which one *selects* variables?

::: {.callout-tip collapse="true"}
## Solution
```{r}
#| label: ex-ridge-lasso
fit_at <- function(a) {
  psolve(Problem(Minimize(loss + elastic_reg(beta, 0.15, a))))
  value(beta)
}
ridge_b <- fit_at(0)
lasso_b <- fit_at(1)
c(ridge_zeros = sum(abs(ridge_b) < 1e-4),
  lasso_zeros = sum(abs(lasso_b) < 1e-4))
```
Ridge shrinks every coefficient but zeroes essentially none; the lasso zeroes the
large majority. That is the defining difference --- only the $\ell_1$ penalty performs
**variable selection**. The elastic net ($0 < \alpha < 1$) blends the two, which helps
when correlated predictors should enter or leave together.
:::

## Takeaways

- Ridge, lasso, and elastic net are **one family**, selected by $\alpha$ --- and in
  CVXR, one small function.
- With a real sparse signal, the lasso **recovers the support** and matches `glmnet`;
  the regularization path is just the **same problem re-solved** over a $\lambda$ grid.
- Only the $\ell_1$ (lasso) part sets coefficients exactly to zero --- the source of
  its variable-selection power.

For more, see [Elastic Net](https://cvxr.rbind.io/examples/regression/elastic-net.html).
