# Changing the Loss: Huber, Quantile, LAD {#sec-change-the-loss}

<!-- LIVE. Source: cvxr_docs/examples/regression/huber-regression.qmd,
     regression/quantile-regression.qmd. Theme: swap the loss, get a different
     estimator. Builds on the LAD exercise in Ch.1 and the `loss` line in Ch.3. -->

::: callout-note
## What you'll be able to do
Swap a squared-error loss for a **Huber** (robust) or **quantile** (asymmetric)
loss by editing a single line, and understand how each changes the estimator.
:::

```{r}
#| label: setup
#| include: false
library(CVXR)
library(ggplot2)
source("_check_status.R")
```

## The loss is a knob

Every regression in this book has the shape

$$\underset{\beta}{\mbox{minimize}}\quad \sum_i \phi\big(y_i - x_i^\top\beta\big),$$

and the choice of the **loss** $\phi$ *is* the choice of estimator. Squared loss
gives least squares. In @sec-gentle-intro we already swapped it for absolute value
($\phi(u)=|u|$) to get least-absolute-deviation regression --- one edited line. This
chapter takes that idea seriously: two more losses, two more estimators, same code.

## Huber: robust to outliers

Least squares is famously sensitive to outliers --- a few bad points drag the whole
fit. The **Huber loss** behaves like squared error for small residuals but only
*linearly* for large ones, so outliers pull less:

$$
\phi_M(u) = \begin{cases} u^2 & |u| \le M \\ 2M|u| - M^2 & |u| > M. \end{cases}
$$

To see it, we generate a one-predictor problem and **flip the sign of 10% of the
responses** --- nasty outliers by construction.

```{r}
#| label: huber-data
set.seed(1289)
n <- 450; M <- 1; pflip <- 0.10
beta_true <- 5
X <- matrix(rnorm(n), ncol = 1)
y_true <- X * beta_true
flip <- 2 * rbinom(n, 1, 1 - pflip) - 1     # -1 flips the sign
y <- flip * y_true + rnorm(n)
```

Now fit the *same* problem with two losses --- squared and Huber:

```{r}
#| label: huber-fit
beta <- Variable(1)

ols <- Problem(Minimize(sum_squares(y - X %*% beta)))
psolve(ols); beta_ols <- value(beta)

hub <- Problem(Minimize(sum(huber(y - X %*% beta, M))))
psolve(hub); beta_hub <- value(beta)

c(true = beta_true, OLS = beta_ols, Huber = beta_hub)
```

The Huber estimate sits far closer to the truth (5); OLS is dragged toward zero by
the flipped points. Only the **loss line changed**.

```{r}
#| label: huber-plot
#| fig-cap: "OLS is pulled toward the outliers; Huber resists."
ggplot(data.frame(X = X, y = y), aes(X, y)) +
  geom_point(alpha = 0.3) +
  geom_abline(aes(slope = beta_ols,  intercept = 0, color = "OLS"),   linewidth = 1) +
  geom_abline(aes(slope = beta_hub,  intercept = 0, color = "Huber"), linewidth = 1) +
  geom_abline(aes(slope = beta_true, intercept = 0, color = "Truth"),
              linetype = "dashed") +
  scale_color_manual(values = c(OLS = "red", Huber = "blue", Truth = "black"),
                     name = NULL) +
  theme_minimal()
```

## Quantile: an asymmetric loss

Least squares estimates the conditional **mean**. To estimate a conditional
**quantile** --- say the 10th or 90th percentile --- we tilt the absolute-value loss
so that over- and under-prediction are penalized unequally:

$$\phi_\tau(u) = \tfrac{1}{2}|u| + \big(\tau - \tfrac{1}{2}\big)u,
\qquad \tau \in (0,1).$$

At $\tau = 0.5$ this is symmetric (the median / LAD); push $\tau$ up or down and the
fit tracks a higher or lower quantile. In CVXR it is, again, a one-line function:

```{r}
#| label: quant-loss
quant_loss <- function(u, tau) 0.5 * abs(u) + (tau - 0.5) * u
```

We make data whose **spread grows with $x$** (heteroscedastic), so different
quantiles fan out:

```{r}
#| label: quant-data
set.seed(7)
nq <- 200
xq <- runif(nq, 0, 4)
yq <- 1 + 2 * xq + rnorm(nq, sd = 0.5 + xq)   # noise widens with x
Xq <- cbind(1, xq)
```

Fit several quantiles --- the *same* solve, only `tau` changes:

```{r}
#| label: quant-fit
taus <- c(0.1, 0.5, 0.9)
bq <- Variable(2)
coefs <- sapply(taus, function(tau) {
  prob <- Problem(Minimize(sum(quant_loss(yq - Xq %*% bq, tau))))
  psolve(prob, solver = "CLARABEL")   # robust choice for the kink at tau = 0.5
  check_solver_status(prob)
  value(bq)
})
colnames(coefs) <- paste0("tau=", taus)
round(coefs, 3)
```

```{r}
#| label: quant-plot
#| fig-cap: "Quantile fits fan out where the spread is largest."
df_lines <- data.frame(tau = factor(taus),
                       intercept = coefs[1, ], slope = coefs[2, ])
ggplot(data.frame(xq, yq), aes(xq, yq)) +
  geom_point(alpha = 0.3) +
  geom_abline(data = df_lines,
              aes(intercept = intercept, slope = slope, color = tau),
              linewidth = 1) +
  labs(x = "x", y = "y", color = expression(tau)) +
  theme_minimal()
```

The three lines spread apart as $x$ (and the noise) grows --- exactly what quantile
regression is for.

## Exercise

**Exercise (⭐).** The Huber threshold $M$ interpolates between two losses. What does
Huber regression become as $M \to 0$? As $M \to \infty$? Check the small-$M$ case
against LAD --- the loss you met in @sec-gentle-intro.

::: {.callout-tip collapse="true"}
## Solution
As $M \to 0$ the quadratic region vanishes and Huber $\to$ (a scaled) **absolute
loss** --- LAD regression. As $M \to \infty$ everything is "small," so Huber $\to$
**squared loss** --- OLS.
```{r}
#| label: ex-huber-M
small_M <- Problem(Minimize(sum(huber(y - X %*% beta, 0.01))))
psolve(small_M); val_small <- value(beta)
lad <- Problem(Minimize(sum(abs(y - X %*% beta))))
psolve(lad); val_lad <- value(beta)
c(huber_smallM = val_small, LAD = val_lad)
```
The two nearly coincide.
:::

## Takeaways

- The **loss** is the estimator: squared → mean, Huber → robust, tilted-$\ell_1$ →
  quantiles, absolute → median.
- Each is a **one-line change** to the objective --- the modeling vocabulary you are
  building transfers directly.

See [Huber](https://cvxr.rbind.io/examples/regression/huber-regression.html) and
[Quantile](https://cvxr.rbind.io/examples/regression/quantile-regression.html).
