# Mixed-Integer Programming {#sec-mip}

<!-- LIVE hands-on: MILP on HiGHS (install-free). DEMO: integer least squares
     via SOC reformulation on scip.
     AUDIT (2026-06-09, verified vs Gurobi): MILP runs on HiGHS & SCIP. A quad
     objective + integers (MIQP) routes through CVXR's QP path -> only commercial
     QP solvers; BUT reformulating the quadratic as an SOC epigraph makes it
     MI-SOCP, which SCIP solves on CRAN to the SAME integer optimum as Gurobi.
     HiGHS silently returns wrong 'infeasible' for SOC-in-MIP -> never send it. -->

::: callout-note
## What you'll be able to do
Add **integer** or **boolean** decision variables to a model and solve the
resulting mixed-integer *linear* program with HiGHS --- no extra install. See a
mixed-integer *second-order-cone* problem solved with SCIP, and learn which
mixed-integer problems still need a commercial solver.
:::

```{r}
#| label: setup
#| include: false
library(CVXR)
```

## Discrete decisions

So far every variable was continuous. But many real decisions are **discrete**:
*select this project or not*, *how many units to ship*, *which facility to open*.
Encoding those as integer or boolean variables turns a convex problem into a
**mixed-integer** one. These are no longer convex --- they are combinatorial and
NP-hard in general --- but CVXR hands them to a solver that does branch-and-bound
over a sequence of convex relaxations. You write the model the same way; you just
declare some variables `integer = TRUE` or `boolean = TRUE`.

## Hands-on: a 0/1 knapsack (MILP)

The classic discrete problem: choose a subset of items to maximize total value
without exceeding a budget. With a **linear** objective and **boolean** variables
this is a mixed-integer *linear* program (MILP) --- squarely within HiGHS, one of
CVXR's built-in CRAN solvers.

```{r}
#| label: knapsack-data
values  <- c(8, 5, 3, 6, 4, 7)   # value of each project
costs   <- c(5, 3, 2, 4, 3, 5)   # its cost
budget  <- 11
projects <- paste0("P", seq_along(values))
```

```{r}
#| label: knapsack-solve
pick <- Variable(length(values), boolean = TRUE)     # 1 = select, 0 = skip
prob <- Problem(Maximize(sum(values * pick)),
                list(sum(costs * pick) <= budget))
psolve(prob, solver = "HIGHS")

selected <- projects[value(pick) > 0.5]
cat("Selected:", paste(selected, collapse = ", "),
    "\nTotal value:", sum(values[value(pick) > 0.5]),
    "  Total cost:", sum(costs[value(pick) > 0.5]), "/", budget, "\n")
```

The `boolean = TRUE` is the whole story --- it forces each `pick` to 0 or 1. Drop it
and you would get the (fractional) **LP relaxation**, an upper bound that is often
not achievable with whole items:

```{r}
#| label: knapsack-relax
pick_lp <- Variable(length(values))
relax <- Problem(Maximize(sum(values * pick_lp)),
                 list(sum(costs * pick_lp) <= budget, pick_lp >= 0, pick_lp <= 1))
cat("LP relaxation value:", round(psolve(relax, solver = "HIGHS"), 3),
    "(>= the integer optimum)\n")
round(value(pick_lp), 3)
```

That gap between the relaxation and the integer optimum is exactly what
branch-and-bound has to close.

## Demo: integer least squares, by reformulation (SCIP)

Here is a mixed-integer problem straight out of our regression chapters: **integer
least squares** --- find *integer* coefficients $x$ minimizing $\|Ax - b\|_2^2$. The
obvious transcription stops us cold:

```{r}
#| label: ils-naive
#| error: true
set.seed(0)
A <- matrix(runif(12 * 5), 12, 5); b <- rnorm(12)
x <- Variable(5, integer = TRUE)
psolve(Problem(Minimize(sum_squares(A %*% x - b))), solver = "SCIP")
```

The message points to commercial solvers --- but that is a **routing** quirk, not a
real limit. A quadratic objective with integers travels through CVXR's *QP path*,
where only GUROBI/CPLEX/XPRESS are registered. The cure is the skill from @sec-dcp:
**reformulate**. Minimizing $\|Ax-b\|_2^2$ has the same minimizer as minimizing the
*norm* $\|Ax-b\|_2$, and a norm is a second-order cone --- which makes this an
**MI-SOCP**, exactly what SCIP handles:

```{r}
#| label: ils-reform
#| eval: !expr requireNamespace("scip", quietly = TRUE)
t <- Variable()
prob <- Problem(Minimize(t), list(p_norm(A %*% x - b, 2) <= t))  # SOC epigraph
psolve(prob, solver = "SCIP")
xhat <- value(x)
cat("integer x:", xhat, "  sum of squares:", round(sum((A %*% xhat - b)^2), 4), "\n")
```

That is the **same integer optimum** a commercial MIQP solver returns --- obtained on
a CRAN solver simply by writing the objective in a form SCIP accepts. SOC plus
integers is SCIP's home turf; the second-order cone is doing the work the quadratic
objective could not.

::: callout-warning
## Two traps worth remembering
- **Never send an SOC-containing mixed-integer problem to HiGHS.** It does not error
  --- it silently returns a *wrong* `infeasible`. SOC + integers needs **SCIP**.
- **A quadratic objective with integers (MIQP)** routes through CVXR's QP path, which
  only commercial solvers serve. On CRAN, **reformulate the quadratic as an SOC**
  (above) and use SCIP --- same answer. (Commercial solvers can take the squared form
  directly, if you have one.)
:::

## Exercise

**Exercise.** Add a rule to the knapsack: projects `P1` and `P2` are mutually
exclusive (at most one may be chosen). Re-solve and see what changes.

::: {.callout-tip collapse="true"}
## Solution
A "select at most one" rule is the linear constraint `pick[1] + pick[2] <= 1`:
```{r}
#| label: ex-knapsack
prob2 <- Problem(Maximize(sum(values * pick)),
                 list(sum(costs * pick) <= budget,
                      pick[1] + pick[2] <= 1))
psolve(prob2, solver = "HIGHS")
projects[value(pick) > 0.5]
```
Logical rules like mutual exclusion, prerequisites (`pick[a] <= pick[b]`), or "pick
exactly k" are all just linear constraints on 0/1 variables.
:::

## Takeaways

- Declaring a variable `integer = TRUE` or `boolean = TRUE` turns a model into a
  mixed-integer program --- same modeling, combinatorial solving.
- **MILP** (linear objective) runs on **HiGHS**, built in and install-free.
- **MI-SOCP** (a cone plus integers) runs on **SCIP**. An integer problem with a
  *quadratic* objective (MIQP) goes to a commercial solver through CVXR's QP path ---
  but **reformulate the quadratic as an SOC** and SCIP solves it on CRAN, same answer.
- Match the solver to the problem --- and never trust an SOC sent to a pure-MILP
  solver like HiGHS.

See [Mixed-Integer QP](https://cvxr.rbind.io/examples/basic/miqp.html) and
[Integer Programming](https://cvxr.rbind.io/examples/misc/integer-programming.html).
