# Logistic Regression {#sec-logistic}

<!-- LIVE. Source: cvxr_docs/examples/regression/logistic-regression.qmd.
     Reinforces the Ch3 DCP/reformulation lesson (logistic atom vs the rejected
     log(1+exp(.))). Matches stats::glm. Verified vs CVXR 1.9.1. -->

::: callout-note
## What you'll be able to do
Fit a binary classifier by maximum likelihood in CVXR, recognize why the natural
expression is *not* DCP and which atom fixes it, match `stats::glm`, and read off
predicted probabilities --- then add a penalty for regularized logistic regression.
:::

```{r}
#| label: setup
#| include: false
library(CVXR)
source("_check_status.R")
```

## From regression to classification

So far our responses were continuous. Now $y_i \in \{0, 1\}$ is a **class label**.
Logistic regression models the probability of class 1 as
$g_\beta(x) = 1/(1 + e^{-x^\top\beta})$ and fits $\beta$ by maximizing the
log-likelihood

$$
\underset{\beta}{\mbox{maximize}}\quad
\sum_{i=1}^m \Big[\, y_i\, x_i^\top\beta - \log\big(1 + e^{x_i^\top\beta}\big) \Big].
$$

The first term is linear; the second is a sum of $\log(1 + e^{z})$ terms.

## A DCP snag --- and the atom that fixes it

You might write that second term directly as `log(1 + exp(X %*% beta))`. CVXR will
**reject it** --- recall @sec-dcp: `exp` is convex, so `1 + exp(...)` is convex, but
`log` is concave-increasing and needs a *concave* argument. The composition is not
verifiable:

```{r}
#| label: nondcp
beta <- Variable(2)
X0 <- matrix(rnorm(20), 10, 2)
is_dcp(Problem(Minimize(sum(log(1 + exp(X0 %*% beta))))))   # FALSE
```

The function *is* convex; DCP just can't see it through that expression. The fix is
the dedicated **`logistic`** atom, $\texttt{logistic}(z) = \log(1 + e^{z})$, which DCP
knows is convex. This is the reformulation habit again --- reach for the atom.

## Fit it

We generate data from a known logistic model and maximize the log-likelihood, writing
$\sum_i \log(1+e^{x_i^\top\beta})$ as `sum(logistic(X %*% beta))`:

```{r}
#| label: fit
set.seed(42)
m <- 500; n <- 8
beta_true <- rnorm(n)
X <- matrix(rnorm(m * n), m, n)
y <- as.numeric(1 / (1 + exp(-X %*% beta_true)) > runif(m))   # 0/1 labels

beta <- Variable(n)
loglik <- sum(y * (X %*% beta)) - sum(logistic(X %*% beta))
prob <- Problem(Maximize(loglik))
psolve(prob)
check_solver_status(prob)
```

It matches `stats::glm` to solver tolerance:

```{r}
#| label: glm
g <- glm(y ~ 0 + X, family = "binomial")
cbind(CVXR = round(value(beta), 4), glm = round(coef(g), 4))
```

## Anything is a function of the solution

A defining CVXR convenience: once solved, you can evaluate *any* expression in the
variables with `value()`. The fitted **probabilities** and **log-odds** need no special
accessor --- just write the formula:

```{r}
#| label: predict
log_odds <- value(X %*% beta)             # the linear predictor
probs    <- 1 / (1 + exp(-log_odds))      # fitted class-1 probabilities
summary(probs)
```

## Regularized logistic regression

Because the objective is just an expression, the regularization idea from
@sec-elastic-net drops straight in. Add an $\ell_1$ penalty for a **sparse** logistic
fit --- one line:

```{r}
#| label: penalized
lambda <- 25
pen_obj <- loglik - lambda * p_norm(beta, 1)   # penalized log-likelihood
psolve(Problem(Maximize(pen_obj)))
round(value(beta), 3)
```

Larger `lambda` drives more coefficients to zero, just as the lasso did for linear
regression --- logistic regression and penalized regression are the *same modeling
vocabulary* applied to a different loss.

## Exercises

**Exercise 1.** Put "anything is a function of the solution" to work. Using the
(unpenalized) fitted model, compute the in-sample **predicted probabilities**, turn
them into class predictions at a 0.5 threshold, and report the classification accuracy
and a confusion matrix.

::: {.callout-tip collapse="true"}
## Solution
Re-solve the maximum-likelihood fit (the penalized sections above changed `beta`), then
everything is just `value()` of an expression:
```{r}
#| label: ex-fitted
psolve(prob)                                  # the unpenalized MLE fit
p_hat <- 1 / (1 + exp(-value(X %*% beta)))    # fitted probabilities
y_hat <- as.numeric(p_hat >= 0.5)             # predicted classes
c(accuracy = mean(y_hat == y))
table(predicted = y_hat, actual = y)
```
The log-odds `X %*% beta` is an *expression*; `value()` evaluates it at the solution,
and the rest is ordinary R. The very same pattern would score *new* data --- just
substitute a new design matrix.
:::

**Exercise 2.** Fit a ridge-penalized ($\ell_2$) logistic regression at the same
`lambda = 25` and compare its coefficients with the $\ell_1$ fit above. Which zeroes
coefficients?

::: {.callout-tip collapse="true"}
## Solution
```{r}
#| label: ex-ridge
l1_fit <- { psolve(Problem(Maximize(loglik - 25 * p_norm(beta, 1))));  value(beta) }
l2_fit <- { psolve(Problem(Maximize(loglik - 25 * sum_squares(beta)))); value(beta) }
data.frame(l1 = round(l1_fit, 3), l2 = round(l2_fit, 3))
```
The $\ell_2$ penalty *shrinks* every coefficient smoothly but zeroes none; the
$\ell_1$ penalty sets several exactly to zero. Same trade-off as ridge vs. lasso in
@sec-elastic-net --- the loss changed, the regularization logic did not.
:::

## Takeaways

- Logistic regression is **maximum likelihood**, a convex problem CVXR solves directly,
  matching `stats::glm`.
- The natural `log(1 + exp(...))` is **non-DCP**; the **`logistic`** atom is the
  DCP-compliant tool --- reformulation, again.
- Predicted probabilities and any other quantity are just `value()` of an expression,
  and penalties from @sec-elastic-net add in one line.

See [Logistic Regression](https://cvxr.rbind.io/examples/regression/logistic-regression.html).
