# Parameters and Efficient Sweeps (DPP) {#sec-dpp}

<!-- LIVE (CORE). Source: cvxr_docs/examples/dpp/tutorial.qmd. The climax of the
     regression spine: turn Ch.3's lambda sweep into a Parameter and reap the
     compilation-caching speedup. -->

::: callout-note
## What you'll be able to do
Replace a numeric constant with a `Parameter`, recognize when a parametrized
problem follows the **DPP** rules, and re-solve it over a grid of values far faster
than rebuilding the problem each time.
:::

```{r}
#| label: setup
#| include: false
library(CVXR)
source("_check_status.R")
```

## A hidden cost in Chapter 3

In @sec-elastic-net we traced a regularization path by building a **new problem**
for each $\lambda$ and solving it. That works, but it hides a cost: every solve
re-runs CVXR's *compilation* --- the translation of your symbolic problem into the
matrices a solver wants. When you sweep dozens or hundreds of values, you pay that
toll over and over.

**Parameters** fix this. A `Parameter` is a symbolic placeholder for a constant
whose value you can change *without rebuilding the problem*. If the problem obeys
the **Disciplined Parametrized Programming (DPP)** rules, CVXR compiles it **once**,
caches the parameter-to-data mapping, and later solves just swap in the new number.

## Parameters and the DPP rule

A `Parameter` looks like a variable but stands for data you will supply:

```{r}
#| label: param
gamma <- Parameter(nonneg = TRUE)   # a nonnegative constant, value set later
x <- Variable(5)
prob <- Problem(Minimize(sum_squares(x - 1) + gamma * p_norm(x, 1)))
is_dpp(prob)
```

DPP adds one main restriction to DCP: **you may not multiply two parametrized
expressions** (a parameter times a parameter, or a parameter times a
parameter-containing expression). Such a problem is still DCP, just not DPP --- so
it solves, but without the speedup:

```{r}
#| label: nondpp
#| results: hold
g <- Parameter(nonneg = TRUE); z <- Variable()
bad <- Problem(Minimize(g * g * z), list(z >= 1))
cat("Is DPP?", is_dpp(bad), "\n")
cat("Is DCP?", is_dcp(bad), "\n")
```

Just like the DCP reformulation skill from @sec-dcp, a non-DPP problem can usually
be **rewritten** to comply --- here by introducing a helper variable:

```{r}
#| label: dpp-fix
#| results: hold
z <- Variable(); w <- Variable(); g <- Parameter(nonneg = TRUE)
good <- Problem(Minimize(g * w), list(w == g * z, z >= 1))
cat("Is DPP?", is_dpp(good), "\n")
```

## The payoff: watch the compile time drop

Recall from @sec-gentle-intro that `verbose = TRUE` splits each solve into a
**compile time** and a **solver time** in its summary. Parameters target the
*compile* time --- so let's just watch it. We set up a lasso whose penalty weight is
a `Parameter`, and solve it the first time:

```{r}
#| label: verbose-first
#| message: true
set.seed(1)
n <- 120; p <- 40
A <- matrix(rnorm(n * p), n, p); b <- rnorm(n)
x <- Variable(p)
error <- sum_squares(A %*% x - b)

gamma <- Parameter(nonneg = TRUE)
param_prob <- Problem(Minimize(error + gamma * p_norm(x, 1)))   # DPP

value(gamma) <- 0.5
psolve(param_prob, verbose = TRUE)        # first solve: pays compilation
```

Look at the **`Compile time`** in the summary. Now change the parameter and re-solve
the *same problem object*:

```{r}
#| label: verbose-resolve
#| message: true
value(gamma) <- 0.7
psolve(param_prob, verbose = TRUE)        # re-solve: compilation is cached
```

The compile time has **dropped** --- CVXR reused the cached parameter-to-data
mapping and went straight to the solver. A freshly built problem, by contrast, pays
the full compile *every* time.

## Quantifying it

Now let's measure the per-solve saving with `bench::mark`, which reports the
garbage-collection-free **minimum** time --- the honest way to compare. We pit one
**DPP re-solve** against one **rebuild** (a fresh problem with a numeric weight):

```{r}
#| label: bench-run
library(bench)
bench::mark(
  `DPP re-solve` = { value(gamma) <- 0.7; psolve(param_prob) },
  `rebuild`      = { psolve(Problem(Minimize(error + 0.7 * p_norm(x, 1)))) },
  check = FALSE, min_iterations = 5, filter_gc = FALSE
)[, c("expression", "min", "median", "mem_alloc")]
```

On our machine the DPP re-solve is roughly **2–3× faster per solve** and allocates
less memory --- precisely the compile time we just watched drop. Over a sweep of
$K$ values you save that difference $K$ times.

::: callout-note
## How big is the win, really?
The saving is exactly the **recompilation you avoid**, so it is largest when
compilation is a meaningful share of each solve --- small-to-medium problems solved
*many* times (regularization paths, grid searches, simulations, bootstraps). For
very large problems the solver itself dominates and the ratio shrinks toward 1×.
Either way the parametrized code is cleaner, and --- as we note below --- parameters
are also the gateway to *differentiating through* a problem.
:::

## Back to the regularization path

This is exactly Chapter 3's sweep, done right. We make $\lambda$ a `Parameter` and
re-solve --- the path looks identical, but only one compilation happens:

```{r}
#| label: dpp-path
lambda_vals <- 10^seq(-3, 1, length.out = 30)
lambda <- Parameter(nonneg = TRUE)
beta <- Variable(p)
path_prob <- Problem(Minimize(sum_squares(A %*% beta - b) + lambda * p_norm(beta, 1)))

betas <- sapply(lambda_vals, function(l) { value(lambda) <- l; psolve(path_prob); value(beta) })
dim(betas)   # p coefficients x number of lambda values
```

Whenever you plan to solve the *same shape* of problem many times --- a path, a
grid search, a simulation, a bootstrap --- reach for a `Parameter`.

## The other half of the solve: warm starts

Recall that every solve splits into a **compile time** and a **solver time**. DPP
mitigates the *compile* time --- it reuses the canonicalized problem structure so we
never re-stuff the matrices. But the solver still starts **cold** each time, refining
an internal guess from scratch until it converges.

A **warm start** addresses the *other* half. A numerical solver is iterative; if you
hand it a starting point already near the answer, it converges in fewer iterations.
In a sweep over a fine grid the solution at one $\lambda$ is an excellent guess for
the next, so the previous answer is exactly the start point we want. In CVXR that is
one flag --- `warm_start = TRUE` uses the current variable values as the solver's
initial point:

```{r}
#| label: warm-sweep
betas_warm <- sapply(lambda_vals, function(l) {
  value(lambda) <- l
  psolve(path_prob, warm_start = TRUE)   # start from the previous lambda's solution
  value(beta)
})
max(abs(betas_warm - betas))             # same path, to solver tolerance
```

The two ideas are **orthogonal and complementary**, not the same trick:

| | reuses | cuts | wins biggest when |
|---|---|---|---|
| **DPP** | the compiled problem structure | *compile* time | compilation dominates --- small problems solved *many* times |
| **Warm start** | the previous solution | *solver* time | the solver dominates --- large/expensive problems, close-together solves |

DPP does nothing for the solver's iterations; a warm start does nothing for
compilation. Each pays off precisely where the other does not, so a tight sweep wants
**both** --- compile once (DPP), then step the solver from each previous answer
(warm start). Warm-starting is solver-dependent (about 9 of CVXR's 15 solvers support
it --- OSQP, SCS, Clarabel, MOSEK, Gurobi, PIQP, HiGHS, and more); a solver that
cannot warm-start simply ignores the flag.

## Exercise

**Exercise (⭐).** The expression `gamma * p_norm(x, 1)` is DPP, but
`(gamma * gamma) * p_norm(x, 1)` is not. Suppose you really want the penalty weight
to be $\gamma^2$. How can you keep the problem DPP *and* sweep over $\gamma$?

::: {.callout-tip collapse="true"}
## Solution
Do the nonlinear part **in R**, outside the DSL: keep a single parameter that holds
the value $\gamma^2$, and set it from your $\gamma$ grid.
```{r}
#| label: ex-dpp
w <- Parameter(nonneg = TRUE)            # will hold gamma^2
prob <- Problem(Minimize(error + w * p_norm(x, 1)))
is_dpp(prob)
for (g in lambda_vals) { value(w) <- g^2; psolve(prob) }   # sweep, still DPP
```
The rule of thumb: parameters may enter *affinely*; any nonlinear transform of a
parameter should be computed in R and stored in another parameter.
:::

## Takeaways

- A **`Parameter`** is a constant you can change without rebuilding the problem.
- If the problem is **DPP** (`is_dpp()` --- essentially "no parameter × parameter"),
  CVXR compiles once and re-solves cheaply.
- Use it for any **repeated solve** --- paths, grid searches, simulations. It is the
  efficient way to do the sweep you met in @sec-elastic-net.
- **DPP cuts compile time; a warm start cuts solver time.** They are independent ---
  combine `Parameter`s with `warm_start = TRUE` to speed up both halves of a sweep.

For more, see the [DPP Tutorial](https://cvxr.rbind.io/examples/dpp/tutorial.html).
