# Isotonic Regression

```{r, echo = FALSE, message = FALSE}
library(isotone)
```

## Goals

- Formulate the isotonic regression objective using some new `CVXR`
  atoms
- Compare with results from `isotone` package
- Exercise on handling ties, the secondary method of `isotone`, using
  `CVXR` atoms
- Exercise on handling ties, the tertiary method of `isotone`, using
  `CVXR` atoms


[Isotonic regression](https://en.wikipedia.org/wiki/Isotonic_regression) is
regression with monotonicity constraints. There are several packages in R
to fit isotonic regression models. In this example, we
consider [`isotone`](https://cran.r-project.org/package=isotone), which
uses a pooled-adjacent-violators algorithm (PAVA) and active set
methods to perform the fit.

## Pituitary Data Example

We will use data from the `isotone` package [@isotone] on the size of
pituitary fissures for 11 subjects between 8 and 14 years of age.

```{r}
data("pituitary")
str(pituitary)
```

Since the size is expected to increase with age, an isotonic fit is
suggested, so we fit using the `isotone` package.

```{r}
res_p <- with(pituitary, gpava(age, size))
```

The `CVXR` formulation expresses this pretty much in the mathematical
form. We define a variable `x` of size `n`, the number of
observations. The objective to be minimized is the least-squares error
(`cvxr_norm`), yet another way of specifying least-squares loss. The
monotonicity is specified using the `diff` function.

### Exercise

Can you explain why `CVXR` provide functions such as `cvxr_norm` and `p_norm`
rather than just plain `pnorm`?

#### Solution

In R, `pnorm` is already defined and refers to the density of the normal
distribution. So we use a new generic `cvxr_norm` or `p_norm` (see
[`CVXR` functions](https://cvxr.rbind.io/cvxr_functions/) to avoid
confusion. The function `cvxr_norm` provides some specialized norms for matrices,
whereas `p_norm` allows one to specify $p$.

```{r}
x_p <- with(pituitary, {
      n <- length(size)
      x <- Variable(n)
      objective <- Minimize(cvxr_norm(size - x, 2))
      constraint <- list(diff(x) >= 0)
      problem <- Problem(objective, constraint)
      result <- solve(problem)
      result$getValue(x)
})
```
As the output below shows, the results are very close.

```{r}
print_matrix(cbind(res_p$x, x_p), col_names = c("isotone", "CVXR"))
```

## Handling Ties

Package `isotone` provides additional methods for handling tied data
besides the default `ties = "primary"` method; `ties = "secondary"`
enforces equality within ties, and `ties = "tertiary"` enforces
monotonicity on the means. (The latter may cause individual fits to be
non-monotonic.)

```{r}
res_s <- with(pituitary, gpava(age, size, ties = "secondary"))
res_t <- with(pituitary, gpava(age, size, ties = "tertiary"))
```

### Exercise

Implement the secondary method of ties using `CVXR` and compare the
results with the `isotone` package.

#### Solution

The secondary method for ties just requires an additional constraint
to enforce equality within tied values; no other modification is
necessary. We do this below by figuring out the tied observation
indices using `base::split` and forcing those `x` values to be equal
(i.e. `diff` == 0).

```{r}
x_s <- with(pituitary, {
    n <- length(size)
    x <- Variable(n)
    objective <- Minimize(p_norm(size - x, 2))
    secondary_constraints <- lapply(base::split(x = seq_len(n),
                                                f = age),
                                    function(i) diff(x[i]) == 0)
    constraint <- c(diff(x) >= 0,
                    secondary_constraints)
    problem <- Problem(objective, constraint)
    solve(problem)$getValue(x)
})
```

Here's the comparison table.

```{r, echo = FALSE}
m <- cbind(res_s$x, x_s)
print_matrix(m, col_names = c("Isotone (S)", "CVXR (S)"))
```

### Exercise

Implement the tertiary method for ties using `CVXR` and compare with
the `isotone` package.

#### Solution

The tertiary method requires computing the block means 
for use in enforcing monotonicity. We call the
[`CVXR::vstack`](https://cvxr.rbind.io/cvxr_functions/) function to
create a single vector of the block means.

Basically, `CVXR::hstack` is the equivalent of `base::cbind` and
`CVXR::vstack` is the equivalent of `base::rbind`.

```{r}
x_t <- with(pituitary, {
    n <- length(size)
    x <- Variable(n)
    objective <- Minimize(p_norm(size - x, 2))
    blocks <- base::split(x = seq_len(n),
                          f = pituitary$age)
    block_means <- lapply(blocks, function(i) {
        v <- numeric(n)
        v[i] <- 1.0 / length(i)
        matrix(v, nrow = 1) %*% x
    })
    block_mean_vector <- do.call(vstack, block_means)
    constraint <- list(diff(block_mean_vector) >= 0)
    problem <- Problem(objective, constraint)
    solve(problem)$getValue(x)
})
```

Here's the comparison table.

```{r, echo = FALSE}
m <- cbind(res_t$x, x_t)
print_matrix(m, col_names = c("Isotone (T)", "CVXR (T)"))
```
