# Further Frontiers: DGP and DQCP {#sec-dgp-dqcp}

<!-- MENTION ONLY (pointer chapter), one tiny verified example each. Source:
     cvxr_docs/examples/dgp/*, dqcp/*. Kept deliberately brief. -->

::: callout-note
## What you'll learn about
Two more "disciplined" dialects beyond DCP --- enough to know they exist and when to
reach for them. Both are covered in depth on the CVXR website.
:::

CVXR's convex grammar (DCP) is not the only disciplined rule set. Two siblings extend
the same idea --- *compose from a library, get a guarantee* --- to other problem
classes. We mention them briefly here; follow the links for the full treatment.

## Disciplined Geometric Programming (DGP)

Some problems are not convex in the variables but become convex after a
**log-log** change of variables --- *geometric programs*, built from positive
variables, monomials, and posynomials. They appear in engineering design, and in
statistics in problems involving products and ratios of positive quantities. You
write them naturally and pass `gp = TRUE`. The classic example maximizes the volume
of a box subject to area limits:

```{r}
#| label: setup
#| include: false
library(CVXR)
```

```{r}
#| label: dgp
x <- Variable(pos = TRUE); y <- Variable(pos = TRUE); z <- Variable(pos = TRUE)
prob <- Problem(Maximize(x * y * z),
                list(2 * (x * z + y * z) <= 10, x * y <= 4))
psolve(prob, gp = TRUE)
round(c(x = value(x), y = value(y), z = value(z)), 3)
```

The objective `x * y * z` multiplies three variables --- not DCP at all --- yet DGP
solves it to global optimality. See the
[DGP Tutorial](https://cvxr.rbind.io/examples/dgp/tutorial.html).

## Disciplined Quasiconvex Programming (DQCP)

A **quasiconvex** function has convex sub-level sets but need not be convex itself
(think of ratios like $\|Ax-b\|/c^\top x$). DCP cannot handle these, but DQCP can,
via a sequence of convex feasibility problems; you pass `qcp = TRUE`. A tiny example
minimizes $(a+1)/\sqrt{a}$, which is quasiconvex on $a > 0$:

```{r}
#| label: dqcp
a <- Variable(pos = TRUE)
prob <- Problem(Minimize((a + 1) / sqrt(a)))
c(is_dqcp = is_dqcp(prob), optimum = psolve(prob, qcp = TRUE), a = value(a))
```

The minimum is $2$ at $a = 1$, found without you having to know the function was
quasiconvex. See the
[DQCP Tutorial](https://cvxr.rbind.io/examples/dqcp/tutorial.html).

## Where this leaves us

You now have the whole map: **DCP** (convex, this book's backbone), **DPP**
(parametrized, for fast re-solves and derivatives), **DGP** (log-log convex),
**DQCP** (quasiconvex), and **DNLP** (smooth nonconvex). Each is the same disciplined
bargain --- follow the grammar, get a solvable problem --- and CVXR checks compliance
for you with the matching `is_*()` predicate. The website's
[examples gallery](https://cvxr.rbind.io) has dozens more worked problems across all
of them.
