# Lasso and Elastic Net

```{r, echo = FALSE, message = FALSE}
library(glmnet)
```

## Goals

- Formulate lasso and elastic net regression models 
- Compare with results from `glmnet` package
- Use loss functions besides squared loss with elastic net penalty

## Regularized Regression

Often in applications, we encounter problems that require
regularization to prevent overfitting, introduce sparsity, facilitate
variable selection, or impose prior distributions on parameters. Two
of the most common regularization functions are the $l_1$-norm and
squared $l_2$-norm, combined in the elastic net regression model
(@elasticnet and @glmnet).

\[
\begin{array}{ll} 
\underset{\beta}{\mbox{minimize}} & \frac{1}{2m}\|y - X\beta\|_2^2 +
\lambda(\frac{1-\alpha}{2}\|\beta\|_2^2 + \alpha\|\beta\|_1). 
\end{array}
\]

Here $\lambda \geq 0$ is the overall regularization weight and
$\alpha \in [0,1]$ controls the relative $l_1$ versus squared $l_2$
penalty. Thus, this model encompasses both ridge ($\alpha = 0$) and
lasso ($\alpha = 1$) regression.

It is convenient to define a function that calculates just the
regularization term given the variable and penalty parameters. This
modular approach will allow us to easily incorporate elastic net
regularization into other regression models as we will see below.

```{r}
#' Define the elastic penalty
#' @param beta the arg min variable
#' @param lambda the penalization parameter
#' @param alpha the elastic net parameter, 0 = ridge, 1 = lasso
elastic_penalty <- function(beta, lambda = 0, alpha = 0) {
    ridge <- (1 - alpha) / 2 * sum_squares(beta)
    lasso <- alpha * cvxr_norm(beta, 1)
    lambda * (lasso + ridge)
}
```

We generate some synthetic sparse data for this example.

```{r}
## Problem data
set.seed(4321)
p <- 10
n <- 500
DENSITY <- 0.25    # Fraction of non-zero beta
beta_true <- matrix(rnorm(p), ncol = 1)
idxs <- sample.int(p, size = floor((1 - DENSITY) * p), replace = FALSE)
beta_true[idxs] <- 0
sigma <- 45
X <- matrix(rnorm(n * p, sd = 5), nrow = n, ncol = p)
eps <- matrix(rnorm(n, sd = sigma), ncol = 1)
Y <- X %*% beta_true + eps
```

We fit the elastic net model for several values of $\lambda$.

```{r}
TRIALS <- 10
beta_vals <- matrix(0, nrow = p, ncol = TRIALS)
lambda_vals <- 10^seq(-2, log10(50), length.out = TRIALS)
```

```{r}
beta <- Variable(p)  
loss <- sum_squares(Y - X %*% beta) / (2 * n)
## Elastic-net regression LASSO
alpha <- 1
beta_vals <- sapply(lambda_vals,
                    function (lambda) {
                        obj <- loss + elastic_penalty(beta, lambda, alpha)
                        prob <- Problem(Minimize(obj))
                        result <- solve(prob)
                        result$getValue(beta)
                    })
```

We can now get a table of the coefficients.

```{r, echo = FALSE}
print_matrix(round(beta_vals, 3),
             row_names = sprintf("$\\beta_{%d}$", seq_len(p)),
             col_names = sprintf("$\\lambda = %.3f$", lambda_vals))
```

We plot the coefficients against the regularization. 

```{r}
plot(0, 0, type = "n", main = "CVXR Regularization Path for Lasso Regression",
     xlab = "Log Lambda", ylab = "Coefficients",
     ylim = c(-1, 2), xlim = c(-4, 4))
matlines(log(lambda_vals), t(beta_vals))
```
  
We then compare with the `glmnet` results.

```{r}
model_net <- glmnet(X, Y, family = "gaussian", alpha = alpha,
                    lambda = lambda_vals,
                    standardize = FALSE,
                    intercept = FALSE,
                    thresh = 1e-8)
## Reverse order to match beta_vals
coef_net <- as.data.frame(as.matrix(coef(model_net)[-1, seq(TRIALS, 1, by = -1)]))
```

```{r, echo = FALSE}
print_matrix(round(coef_net, 3),
             row_names = sprintf("$\\beta_{%d}$", seq_len(p)),
             col_names = sprintf("$\\lambda = %.3f$", lambda_vals))
```

### Exercise

[A Stack Overflow
Question](https://stats.stackexchange.com/questions/334007/elastic-net-in-glmnet-vs-cvxr)!

_I'm attempting to match some simple results in R using `glmnet` and
`CVXR`. I have the following code._
```{r}
library(glmnet)
data(QuickStartExample)
catn <- function(...) cat(..., "\n")
objective_value <- function(y, x, coefs, lambda, alpha) {
    n <- nrow(x)
    ridge <- sum(coefs^2) ; l1 <- sum(abs(coefs))
    sum((y - (x %*% coefs))^2) / (2 * n) + lambda * ((1 - alpha) / 2 * ridge + alpha * l1)
}
alpha  <- 0; lambda  <- 1;
fit <- glmnet(x, y, intercept=F, standardize=F, lambda=1, alpha=0)
```
_which gives me one set of coefficients and the objective function
value `2.559086` via_

```{r}
objective_value(y, x, coef(fit)[-1, ], lambda, alpha)
```
_but this `CVXR` code _
```{r}
beta <- Variable(20)
elastic_reg <- function(beta, lambda = 0, alpha = 0) {
   ridge <- (1 - alpha) * sum(beta^2) * .5
   lasso <- alpha * p_norm(beta, 1)
   lambda * (lasso + ridge)
}
loss <- sum((y - x %*% beta)^2)/(2*length(y))
obj <- loss + elastic_reg(beta, lambda = 1, 0)
prob <- Problem(Minimize(obj))
result <- solve(prob)
print(result$value)
```
_gives a different objective value `2.859259` and a somewhat different set of
coefficients. Can you help?_

#### Solution

To compare glmnet results to `CVXR` for the Gaussian case, it is
advisable to standardize the response per the `glmnet` documentation.
Note also that `glmnet` uses $n$ rather than $n-1$ in the denominator
for $y$. This will ensure that the $\lambda$ is on the same scale as
shown below.

```{r}
catn <- function(...) cat(..., "\n")
## Standardize the y
y_s <- local({
    n <- length(y)
    m <- mean(y); s <- as.numeric(sqrt(var(y) * (n - 1) / n));
    result <- (y - m) / s ## scale using 1/n
    attr(result, "scaled:center") <- m
    attr(result, "scaled:scale") <- s
    result
})
```
We can do a comparison on the standardized $y$. First, the `glmnet` answer:

```{r}
## STANDARDIZED COMPARISON
fit_s <- glmnet(x, y_s, intercept=F, standardize=F, lambda = lambda, alpha=alpha)
catn("Glmnet objective (scaled y)",
     objective_value(y_s, x, coef(fit_s)[-1], lambda, alpha))
```

Next, the `CVXR` answer:

```{r}
elastic_reg <- function(beta, lambda = 0, alpha = 0) {
    ridge <- (1 - alpha) / 2 * sum_squares(beta)
    lasso <- alpha * p_norm(beta, 1)
    lambda * (lasso + ridge)
}
loss <- sum_squares(y_s - x %*% beta) / ( 2 * nrow(x))
obj <- loss + elastic_reg(beta, lambda = lambda, alpha)
prob <- Problem(Minimize(obj))
beta_est <- solve(prob)$getValue(beta)
catn("CVXR objective (scaled y):", objective_value(y_s, x, beta_est, lambda, alpha))
```

To work on the non-standardized scale, we need to match the `lamba`
values as noted by the `glmnet` authors in Appendix 2 of the package
vignette [@glmnet].

```{r}
## NONSTANDARDIZED COMPARISON
fit <- glmnet(x, y, intercept=F, standardize=F, lambda = lambda, alpha=alpha)
catn("Glmnet objective (unscaled y)", objective_value(y, x, coef(fit)[-1], lambda, alpha))
```

```{r}
obj <- loss + elastic_reg(beta, lambda = lambda / attr(y_s, "scaled:scale"), alpha)
prob <- Problem(Minimize(obj))
beta_est <- solve(prob)$getValue(beta)
catn("CVXR objective (unscaled y)", objective_value(y, x, beta_est, lambda, alpha))
```

Finally, we can check that the coefficients are close enough.

```{r}
print_matrix(round(cbind(beta_est, coef(fit)[-1]), 3), 
             row_names = sprintf("$\\beta_{%d}$", seq_len(20)),
             col_names = c("CVXR", "GLMNET")) 
```

### Exercise

Using the data (`X`, `Y`) above, solve an elastic net problem with
[Huber loss](https://cvxr.rbind.io/cvxr_functions/) using the Huber
threshold $M = 0.5$.

#### Solution 

Just set the loss as follows.

```{r}
beta <- Variable(p)  
loss  <- huber(Y - X %*% beta, M = 1)
loss <- sum_squares(Y - X %*% beta) / (2 * n)
## Elastic-net regression LASSO
alpha <- 1
beta_vals <- sapply(lambda_vals,
                    function (lambda) {
                        obj <- loss + elastic_penalty(beta, lambda, alpha)
                        prob <- Problem(Minimize(obj))
                        result <- solve(prob)
                        result$getValue(beta)
                    })
```

The estimates are below and don't change in this example.

```{r, echo = FALSE}
print_matrix(round(beta_vals, 3),
             row_names = sprintf("$\\beta_{%d}$", seq_len(p)),
             col_names = sprintf("$\\lambda = %.3f$", lambda_vals))
```
