# Portfolio Optimization {#sec-portfolio}

<!-- LIVE. Mean-variance (Markowitz) efficient frontier + mean-CVaR, on a real
     12-asset price subset bundled in data/stock_prices.csv. Example and data
     adapted from Riskfolio-Lib Tutorial 1 (Dany Cajas). Originals: Markowitz
     (1952) mean-variance; Rockafellar-Uryasev (2000) CVaR. Verified vs CVXR 1.9.1. -->

::: callout-note
## What you'll be able to do
Build a **mean-variance** (Markowitz) portfolio, trace its risk-return **efficient
frontier**, and re-cast the problem with a different risk measure --- **Conditional
Value-at-Risk (CVaR)** --- all from real return data.
:::

::: callout-tip
## Credits
The example and the bundled price data are adapted from
[**Riskfolio-Lib**](https://github.com/dcajasn/Riskfolio-Lib) (Dany Cajas) ---
"Tutorial 1: Classic Mean-Risk Optimization" [@riskfolio], itself built on CVXPY. The
underlying models are due to @Markowitz:1952 (mean-variance) and
@RockafellarUryasev:2000 (CVaR).
:::

```{r}
#| label: setup
#| include: false
library(CVXR)
library(ggplot2)
library(RColorBrewer)
source("_check_status.R")
```

## The mean-variance problem

We hold $n$ assets; $w_i$ is the fraction of our budget in asset $i$. Returns $r$
have mean $\mu$ and covariance $\Sigma$, so a portfolio $w$ has expected return
$\mu^\top w$ and variance (risk) $w^\top\Sigma w$. Markowitz's insight: trade return
against risk with a **risk-aversion** knob $\gamma \ge 0$,

$$
\underset{w}{\mbox{maximize}}\ \ \mu^\top w - \gamma\, w^\top\Sigma w
\qquad\text{subject to}\qquad w \ge 0,\ \ \sum_i w_i = 1.
$$

The constraints make it a *long-only, fully-invested* portfolio. Small $\gamma$ chases
return; large $\gamma$ avoids risk.

## Real data

We use ~3 years of daily prices for 12 well-known stocks (a subset of the Riskfolio-Lib
dataset), and turn prices into simple returns, a mean vector, and a covariance matrix:

```{r}
#| label: data
prices <- read.csv("data/stock_prices.csv", check.names = FALSE)
P <- as.matrix(prices[, -1])              # drop the Dates column
R <- P[-1, ] / P[-nrow(P), ] - 1          # daily simple returns
mu    <- colMeans(R)
Sigma <- cov(R)
n     <- ncol(R)
colnames(R)
```

## Trace the efficient frontier

For each $\gamma$ we solve the Markowitz problem and record the realized risk and
return. This is the **same problem re-solved** over a grid --- exactly where the
`Parameter` trick from @sec-dpp pays off. We make $\gamma$ a `Parameter`, build the
problem **once**, and each iteration just sets `value(gamma)` and re-solves, reusing
the cached compilation instead of canonicalizing from scratch 40 times.

```{r}
#| label: frontier
w     <- Variable(n)
ret   <- t(mu) %*% w
risk  <- quad_form(w, Sigma)              # w' Sigma w, a convex quadratic
cons  <- list(w >= 0, sum(w) == 1)
gamma <- Parameter(nonneg = TRUE)         # the risk-aversion knob, set per solve
prob  <- Problem(Maximize(ret - gamma * risk), cons)   # built once

gammas <- 10^seq(-1, 3, length.out = 40)
fr <- sapply(gammas, function(g) {
  value(gamma) <- g                       # swap in the new gamma; no rebuild
  psolve(prob)
  c(risk = sqrt(value(risk)), ret = value(ret), w = value(w))
})
```

Note how we read the risk and return back by **directly evaluating** the expressions
`sqrt(risk)` and `ret` after each solve --- no bookkeeping required. The frontier:

```{r}
#| label: frontier-plot
#| fig-cap: "Efficient frontier. Red points are single-asset portfolios; the curve dominates them."
df  <- data.frame(risk = fr["risk", ], ret = fr["ret", ])
ind <- data.frame(risk = sqrt(diag(Sigma)), ret = mu, asset = names(mu))
ggplot(df, aes(risk, ret)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(data = ind, color = "red") +
  geom_text(data = ind, aes(label = asset), size = 3, vjust = -0.6,
            check_overlap = TRUE) +
  labs(x = "Risk (standard deviation of daily return)", y = "Expected daily return") +
  theme_minimal()
```

Every point on the blue curve is the best return achievable at that level of risk; the
individual stocks (red) all sit *below* it --- diversification at work.

The frontier shows the risk-return *outcome*, but not *what the portfolio holds*. Since
`fr` already captured the weight vector at every $\gamma$, we can read the composition
straight back out and watch diversification happen. At a few points along the grid we
stack each asset's share of the budget:

```{r}
#| label: frontier-composition
#| fig-cap: "Portfolio composition along the frontier. Small gamma concentrates the budget in a few assets; large gamma spreads it out --- diversification, made visible."
markers <- c(1, 10, 20, 30, 40)
wmat <- fr[grep("^w", rownames(fr)), markers]   # 12 assets x 5 values of gamma
rownames(wmat) <- names(mu)

comp <- data.frame(
  gamma  = factor(rep(signif(gammas[markers], 2), each = n),
                  levels = signif(gammas[markers], 2)),
  asset  = factor(rep(names(mu), times = length(markers)), levels = names(mu)),
  weight = as.vector(wmat))

ggplot(comp, aes(gamma, weight, fill = asset)) +
  geom_col() +
  scale_fill_manual(values = brewer.pal(n, "Paired")) +
  labs(x = expression("Risk aversion  " * gamma),
       y = "Fraction of budget", fill = "Asset") +
  theme_minimal()
```

The left-most bar (greedy, low $\gamma$) leans on one or two names; as $\gamma$ grows the
bars fragment into the full palette. That shift *is* the efficient frontier, seen from
the portfolio's side.

## A different risk measure: CVaR

Variance penalizes upside and downside alike. **Conditional Value-at-Risk** (CVaR)
focuses on the *bad tail*: $\text{CVaR}_\alpha$ is (roughly) the average loss in the
worst $1-\alpha$ fraction of scenarios. Rockafellar and Uryasev showed it has a
beautifully simple convex (in fact linear) form using the historical return scenarios
$r_t$ directly. With losses $L_t = -r_t^\top w$,

$$
\text{CVaR}_\alpha(w) = \min_{\tau}\ \ \tau + \frac{1}{(1-\alpha)T}\sum_{t=1}^{T}\big(L_t - \tau\big)_+ .
$$

In CVXR we introduce an auxiliary `VaR` variable $\tau$ and minimize CVaR subject to a
return target --- a clean linear program:

```{r}
#| label: cvar
alpha  <- 0.95
target <- mean(mu)                        # require at least the average asset's return
Tn     <- nrow(R)

wv  <- Variable(n)
VaR <- Variable(1)
loss <- -R %*% wv                         # vector of T scenario losses
cvar <- VaR + (1 / ((1 - alpha) * Tn)) * sum(pos(loss - VaR))

prob_cvar <- Problem(Minimize(cvar),
                     list(sum(wv) == 1, wv >= 0, t(mu) %*% wv >= target))
psolve(prob_cvar)
check_solver_status(prob_cvar)

w_cvar <- setNames(round(value(wv), 3), names(mu))
w_cvar[w_cvar > 0.001]                     # the minimum-CVaR allocation
```

Same data, same `CVXR` machinery, a different notion of risk --- and a different,
typically more tail-aware, allocation. Switching risk measures is, once again, just a
change to the objective.

To see *how* the allocation differs, hold the return target fixed and put the two
side by side. The fair mean-variance counterpart minimizes variance subject to the same
return floor, so both bars sit at the same expected return and only the risk measure
changes:

```{r}
#| label: cvar-vs-mv
#| fig-cap: "Same return target, two risk measures. Variance and CVaR reallocate the budget differently."
## mean-variance allocation at the SAME return target, for an apples-to-apples comparison
wv_mv   <- Variable(n)
prob_mv <- Problem(Minimize(quad_form(wv_mv, Sigma)),
                   list(sum(wv_mv) == 1, wv_mv >= 0, t(mu) %*% wv_mv >= target))
psolve(prob_mv)
check_solver_status(prob_mv)

comp2 <- data.frame(
  measure = factor(rep(c("Mean-variance", "Mean-CVaR"), each = n),
                   levels = c("Mean-variance", "Mean-CVaR")),
  asset   = factor(rep(names(mu), 2), levels = names(mu)),
  weight  = c(value(wv_mv), value(wv)))

ggplot(comp2, aes(measure, weight, fill = asset)) +
  geom_col() +
  scale_fill_manual(values = brewer.pal(n, "Paired")) +
  labs(x = NULL, y = "Fraction of budget", fill = "Asset") +
  theme_minimal()
```

Both portfolios earn the same expected return, yet CVaR --- caring only about the bad
tail --- tilts the budget toward a different mix than variance, which penalizes upside
and downside alike.

## Exercise

**Exercise (⭐).** Add a **diversification** rule to the mean-variance problem: no
single asset may exceed 25% of the budget. Re-solve at $\gamma = 10$ and compare with
the unconstrained allocation.

::: {.callout-tip collapse="true"}
## Solution
A position cap is one linear constraint, `w <= 0.25`:
```{r}
#| label: ex-cap
base <- Problem(Maximize(ret - 10 * risk), cons)
capd <- Problem(Maximize(ret - 10 * risk), c(cons, list(w <= 0.25)))
psolve(base); w0 <- value(w)
psolve(capd); w1 <- value(w)
data.frame(asset = names(mu),
           uncapped = round(w0, 3),
           capped   = round(w1, 3))[round(w0,3) > 0 | round(w1,3) > 0, ]
```
The cap spreads weight off the assets that the unconstrained solution concentrated in
--- a one-line way to encode an investment policy.
:::

## Takeaways

- **Mean-variance** optimization is a convex quadratic program; sweeping the
  risk-aversion $\gamma$ traces the **efficient frontier** (a `Parameter` sweep, @sec-dpp).
- Read risk and return straight off the solved expressions with `value()`.
- Swapping variance for **CVaR** is just a new objective --- CVXR makes alternative risk
  measures, position limits, and policy constraints one-liners.

For many more portfolio models (risk parity, Black-Litterman, worst-case, drawdown
measures), see [Riskfolio-Lib](https://github.com/dcajasn/Riskfolio-Lib) and the
[CVXR portfolio example](https://cvxr.rbind.io/examples/finance/portfolio-optimization.html).
