# Sparse Inverse Covariance Estimation {#sec-invcov}

<!-- LIVE. Adapted from cvxr_docs statistics/sparse-inverse-covariance-estimation.qmd
     (Fu & Narasimhan). Sampling uses base chol() instead of expm::sqrtm to keep
     dependencies light. Verified vs CVXR 1.9.1 (2026-06-10): true nonzeros = 24;
     alpha=10 -> 46, alpha=4 -> 24 (recovers truth), alpha=1 -> 12. -->

::: callout-note
## What you'll be able to do
Estimate the **graph** of conditional dependencies among variables --- which ones are
related once you account for all the others --- by fitting a **sparse inverse
covariance** matrix with a log-determinant objective and an $\ell_1$ budget.
:::

```{r}
#| label: setup
#| include: false
library(CVXR)
library(Matrix)
library(ggplot2)
source("_check_status.R")
```

## The idea

The portfolio chapter (@sec-portfolio) took a covariance matrix as given. Here we go
the other way and **estimate** it --- and specifically its inverse, which carries the
more interesting story.

For Gaussian data, the **inverse** covariance $S = \Sigma^{-1}$ (the *precision*
matrix) encodes **conditional independence**: $S_{ij} = 0$ exactly when variables $i$
and $j$ are independent *given all the others*. So a **sparse** $S$ is a sparse graph
of direct relationships --- often what we actually want to discover. With more
variables than the data can pin down, we impose that sparsity directly.

Given i.i.d. samples $x_i \sim N(0, \Sigma)$ with sample covariance $Q$, we maximize
the Gaussian log-likelihood while capping the total size of $S$:

$$
\begin{array}{ll}
\underset{S}{\mbox{maximize}} & \log\det(S) - \mathop{\mathrm{tr}}(SQ) \\
\mbox{subject to} & S \succeq 0, \qquad \sum_{i,j} |S_{ij}| \le \alpha .
\end{array}
$$

The $\ell_1$ budget $\alpha \ge 0$ controls sparsity: smaller $\alpha$ forces more
off-diagonal entries to exactly zero. The objective is concave (`log_det` is concave;
the trace term is linear) and the constraints are convex, so CVXR solves it directly.

## Synthetic data with a known answer

So we can check the estimate, we build a *true* sparse precision matrix $S_{\text{true}}$,
invert it to get the covariance $R$, and sample from $N(0, R)$:

```{r}
#| label: data
set.seed(1)
p <- 10        # number of variables
n <- 1000      # number of samples

A      <- rsparsematrix(p, p, 0.15, rand.x = rnorm)
Strue  <- as.matrix(A %*% t(A) + 0.05 * diag(p))   # sparse, symmetric, positive definite
R      <- solve(Strue)                              # the covariance to sample from

x <- matrix(rnorm(n * p), n, p) %*% chol(R)         # rows are i.i.d. N(0, R)
Q <- cov(x)                                          # sample covariance

sum(abs(Strue) > 1e-4)                               # how many true nonzeros (of 100)
```

## Fit at several sparsity budgets

A semidefinite variable is declared with `PSD = TRUE`. We sweep $\alpha$ and, after
each solve, zero out numerically negligible entries to read off the recovered pattern:

```{r}
#| label: fit
alphas <- c(10, 6, 4, 1)
S      <- Variable(c(p, p), PSD = TRUE)
obj    <- Maximize(log_det(S) - matrix_trace(S %*% Q))

S_est <- lapply(alphas, function(a) {
  prob <- Problem(obj, list(sum(abs(S)) <= a))
  psolve(prob)
  check_solver_status(prob)
  Sa <- value(S)
  Sa[abs(Sa) <= 1e-4] <- 0            # treat tiny entries as exact zeros
  Sa
})
sapply(S_est, function(M) sum(M != 0))  # nonzeros recovered at each alpha
```

Note `matrix_trace(S %*% Q)` rather than `sum(diag(S %*% Q))` --- same value, but the
dedicated atom keeps the expression tree compact. And `log_det(S)` is its own atom:
there is no `det()` for a CVXR variable, because the log-determinant is the convex
object DCP understands.

## What the budget buys

The sparsity pattern of the estimate tightens as $\alpha$ shrinks. We plot the nonzero
mask of the truth alongside each fit:

```{r}
#| label: pattern-plot
#| fig-width: 9
#| fig-height: 2.4
#| fig-cap: "Nonzero pattern (dark = nonzero) of the true precision matrix and the estimates. Smaller alpha -> sparser. Around alpha = 4 the estimate matches the truth's sparsity."
mask_df <- function(M, lab) {
  data.frame(i = rep(seq_len(p), p), j = rep(seq_len(p), each = p),
             nz = factor(as.integer(as.vector(M) != 0)), panel = lab)
}
panels <- rbind(
  mask_df(Strue, "truth"),
  do.call(rbind, Map(function(M, a) mask_df(M, paste0("alpha = ", a)), S_est, alphas)))
panels$panel <- factor(panels$panel, levels = c("truth", paste0("alpha = ", alphas)))

ggplot(panels, aes(i, j, fill = nz)) +
  geom_tile(color = "grey80") +
  scale_fill_manual(values = c("0" = "white", "1" = "steelblue"), guide = "none") +
  scale_y_reverse() + coord_equal() +
  facet_wrap(~ panel, nrow = 1) +
  theme_void() + theme(strip.text = element_text(margin = margin(4, 0, 4, 0)))
```

At $\alpha = 10$ the budget is loose and the estimate is dense; as $\alpha$ falls the
off-diagonal entries are squeezed to zero, and near $\alpha = 4$ the recovered pattern
matches the true number of nonzeros. Too small ($\alpha = 1$) and we lose real edges.
In practice you would pick $\alpha$ by cross-validation --- and because the only thing
changing across the sweep is one constant, this is a textbook case for the `Parameter`
trick of @sec-dpp.

## Takeaways

- The **inverse** covariance is the object of interest: its zeros are **conditional
  independencies** --- a graph of direct relationships.
- `log_det(S) - matrix_trace(S %*% Q)` with `S` declared `PSD = TRUE` and an
  $\ell_1$ budget is the whole model --- a few lines of CVXR.
- One knob, $\alpha$, trades density for sparsity; sweep it with a `Parameter`.

See [Sparse Inverse Covariance Estimation](https://cvxr.rbind.io/examples/statistics/sparse-inverse-covariance-estimation.html)
for the full treatment.
