# Nearly Isotonic Fits

## Goals

- Formulate nearly-isotonic and nearly-convex fits using `CVXR` atoms
- Use the bootstrap to estimate variance of approximation

```{r, message = FALSE, echo = FALSE}
library(ggplot2)
library(boot)
```

Given a set of data points $y \in {\mathbf R}^m$,
@TibshiraniHoefling:2011 fit a nearly-isotonic approximation $\beta
\in {\mathbf R}^m$ by solving

$$
\begin{array}{ll}
\underset{\beta}{\mbox{minimize}} & \frac{1}{2}\sum_{i=1}^m (y_i - \beta_i)^2 + \lambda \sum_{i=1}^{m-1}(\beta_i - \beta_{i+1})_+,
\end{array}
$$

where $\lambda \geq 0$ is a penalty parameter and $x_+
=\max(x,0)$. This can be directly formulated in `CVXR`.

## Global Warming Example

As an
example, we use global warming data from
the
[Carbon Dioxide Information Analysis Center (CDIAC)](http://cdiac.ess-dive.lbl.gov/ftp/trends/temp/jonescru/). The
data points are the annual temperature anomalies relative to the
1961--1990 mean.

```{r}
data(cdiac)
str(cdiac)
```

Since we plan to fit the regression and also get some idea of the
standard errors, we write a function that computes the fit for use in
bootstrapping. 

```{r}
neariso_fit <- function(y, lambda) {
    m <- length(y)
    beta <- Variable(m)
    A <- as(m:1, "pMatrix") %*% Matrix::Diagonal(m)   # Matrix for reversing order
    obj <- 0.5 * sum_squares(y - beta) + lambda * sum(pos(diff(A %*% beta)))
    prob <- Problem(Minimize(obj))
    solve(prob)$getValue(beta)
}
```

Observe that we reverse the order of `beta` since `diff` outputs 
`beta[i+1] - beta[i]` for $i = 1,\ldots,m$. The `CVXR::pos` atom 
evaluates $x_+ = \max(x,0)$ elementwise on the input expression.

The `boot` library provides all the tools for bootstrapping, but
requires a statistic function that takes particular arguments: a data
frame, followed by the bootstrap indices and any other arguments
($\lambda$ for instance). This is defined below.

_NOTE_ In what follows, we use a very small number of bootstrap
samples as the fits are time consuming.

```{r}
neariso_fit_stat <- function(data, index, lambda) {
    sample <- data[index,]                  # Bootstrap sample of rows
    sample <- sample[order(sample$year),]   # Order ascending by year
    neariso_fit(sample$annual, lambda)
}
```

```{r, eval = FALSE}
set.seed(123)
boot.neariso <- boot(data = cdiac,
                     statistic = neariso_fit_stat,
                     R = 10, lambda = 0.44)
ci.neariso <- t(sapply(seq_len(nrow(cdiac)),
                       function(i) boot.ci(boot.out = boot.neariso, conf = 0.95,
                                           type = "norm", index = i)$normal[-1]))
data.neariso <- data.frame(year = cdiac$year,
                           annual = cdiac$annual,
                           est = boot.neariso$t0,
                           lower = ci.neariso[, 1],
                           upper = ci.neariso[, 2])
```

We can now plot the fit and confidence bands for the nearly-isotonic
fit. 

```{r, eval = FALSE}
(plot.neariso <- ggplot(data = data.neariso) +
     geom_point(mapping = aes(year, annual), color = "red") +
     geom_line(mapping = aes(year, est), color = "blue") +
     geom_ribbon(mapping = aes(x = year, ymin = lower, ymax = upper), alpha = 0.3) +
     labs(x = "Year", y = "Temperature Anomalies")
)
```
The curve follows the data well, but exhibits some choppiness in
regions with a steep trend. 

### Exercise

Fit a smoother curve using a nearly-convex fit described in the same
paper:

$$
\begin{array}{ll}
\underset{\beta}{\mbox{minimize}} & \frac{1}{2}\sum_{i=1}^m (y_i -
\beta_i)^2 + \lambda \sum_{i=1}^{m-2}(\beta_i - 2\beta_{i+1} + \beta_{i+2})_+ \end{array} 
$$

#### Solution

This replaces the first difference term with an approximation to the
second derivative at $\beta_{i+1}$. In `CVXR`, the only change
necessary is the penalty line: replace `diff(A %*% x)` by 
`diff(A %*% x, differences = 2)`.

```{r, eval = FALSE}
nearconvex_fit <- function(y, lambda) {
    m <- length(y)
    beta <- Variable(m)
    A <- as(m:1, "pMatrix") %*% Matrix::Diagonal(m)   # Matrix for reversing order
    obj <- 0.5 * sum_squares(y - beta) + lambda * sum(pos(diff(A %*% beta, differences = 2)))
    prob <- Problem(Minimize(obj))
    solve(prob)$getValue(beta)
}

nearconvex_fit_stat <- function(data, index, lambda) {
    sample <- data[index,]                  # Bootstrap sample of rows
    sample <- sample[order(sample$year),]   # Order ascending by year
    nearconvex_fit(sample$annual, lambda)
}

set.seed(987)
boot.nearconvex <- boot(data = cdiac,
                        statistic = nearconvex_fit_stat,
                        R = 5,
                        lambda = 0.44)

ci.nearconvex <- t(sapply(seq_len(nrow(cdiac)),
                          function(i) boot.ci(boot.out = boot.nearconvex, conf = 0.95,
                                              type = "norm", index = i)$normal[-1]))
data.nearconvex <- data.frame(year = cdiac$year,
                              annual = cdiac$annual,
                              est = boot.nearconvex$t0,
                              lower = ci.nearconvex[, 1],
                              upper = ci.nearconvex[, 2])

```

The resulting curve for the nearly-convex fit is depicted below with
95\% confidence bands generated from $R = 5$ samples. Note the jagged
staircase pattern has been smoothed out. 


```{r, eval = FALSE}
(plot.nearconvex <- ggplot(data = data.nearconvex) +
     geom_point(mapping = aes(year, annual), color = "red") +
     geom_line(mapping = aes(year, est), color = "blue") +
     geom_ribbon(mapping = aes(x = year, ymin = lower, ymax = upper), alpha = 0.3) +
     labs(x = "Year", y = "Temperature Anomalies")
)
```

## References
