# Showcase: Image Reconstruction and the Bootstrap {#sec-showcase}

<!-- LIVE. Source: tv-inpainting.qmd (images bundled in images/) +
     near-isotonic-and-near-convex-regression.qmd (cdiac ships with CVXR).
     Images shown with base rasterImage (no gridExtra dependency). -->

::: callout-note
## What you'll be able to do
Reconstruct a corrupted image by total-variation in-painting, and attach bootstrap
confidence bands to a CVXR fit --- two showcases of convex optimization doing real
work on real data.
:::

```{r}
#| label: setup
#| include: false
library(CVXR)
library(png)
library(ggplot2)
library(boot)
source("_check_status.R")
```

## Repairing a damaged image

A grayscale image is just a matrix of pixel intensities. Suppose some pixels are
corrupted (here, white text scrawled over the picture) and we know *which* ones.
Can we fill them in? The **total-variation in-painting** model says: among all images
that agree with the *known* pixels, choose the one with the smallest total variation
--- the least "jumpy" one. Total variation of $U$ is

$$\mathop{\bf tv}(U) = \sum_{i,j}\left\|\begin{pmatrix} U_{i+1,j}-U_{ij}\\ U_{i,j+1}-U_{ij}\end{pmatrix}\right\|_2,$$

a convex function, so the whole problem is convex. We load a $128\times128$ image and
a corrupted copy, and mark a pixel *known* where the two agree:

```{r}
#| label: load-img
u_orig <- readPNG("images/loki128.png")
u_corr <- readPNG("images/loki128_corrupted.png")
if (length(dim(u_orig)) == 3) u_orig <- (u_orig[,,1] + u_orig[,,2] + u_orig[,,3]) / 3
if (length(dim(u_corr)) == 3) u_corr <- (u_corr[,,1] + u_corr[,,2] + u_corr[,,3]) / 3
rows <- nrow(u_orig); cols <- ncol(u_orig)
known <- 1 * (u_orig == u_corr)        # 1 where the pixel survived
cat(sprintf("%d x %d image; %.0f%% of pixels known\n",
            rows, cols, 100 * sum(known) / (rows * cols)))
```

The model is a direct transcription of the math: minimize total variation subject to
matching the known pixels.

```{r}
#| label: tv-solve
U <- Variable(c(rows, cols))
prob <- Problem(Minimize(total_variation(U)),
                list(known * U == known * u_corr))
psolve(prob, solver = "SCS")          # ~16k variables; a few seconds
U_val <- pmin(pmax(value(U), 0), 1)   # clip to [0, 1]
cat("status:", status(prob), "\n")
```

```{r}
#| label: tv-plot
#| fig-width: 9
#| fig-height: 3.4
#| fig-cap: "Original, corrupted, and in-painted (the text is gone)."
show_img <- function(m, title) {
  plot(c(0, 1), c(0, 1), type = "n", axes = FALSE, xlab = "", ylab = "",
       main = title, asp = 1)
  rasterImage(as.raster(m), 0, 0, 1, 1, interpolate = FALSE)
}
op <- par(mfrow = c(1, 3), mar = c(1, 1, 2, 1))
show_img(u_orig, "Original")
show_img(u_corr, "Corrupted")
show_img(U_val,  "In-painted")
par(op)
```

The solver never "knew" it was looking at a face --- it just found the smoothest
image consistent with the surviving pixels. That is the power of a good convex model.

## Bootstrap confidence bands on a CVXR fit

Our second showcase connects optimization back to statistical **inference**. CVXR
fits often have no closed form, so there is no textbook standard-error formula --- but
the **bootstrap** does not need one. We use global-mean temperature anomalies
(`cdiac`, shipped with CVXR) and a **nearly-isotonic** fit, which is least squares
plus a penalty on downward steps (it prefers a mostly-increasing curve):

$$\underset{\beta}{\mbox{minimize}}\quad \tfrac12\sum_i (y_i-\beta_i)^2
   + \lambda \sum_i (\beta_{i+1}-\beta_i)_+ .$$

```{r}
#| label: neariso-fn
data(cdiac)
neariso_fit <- function(y, lambda) {
  beta <- Variable(length(y))
  prob <- Problem(Minimize(0.5 * sum((y - beta)^2) + lambda * sum(pos(diff(beta)))))
  psolve(prob)
  value(beta)
}
```

To get standard errors we wrap that fit in `boot`. The statistic refits on each
resample; we use a small number of replicates since each one solves an optimization
problem:

```{r}
#| label: neariso-boot
fit_stat <- function(data, idx, lambda) {
  s <- data[idx, ]
  s <- s[order(s$year), ]
  neariso_fit(s$annual, lambda)
}

set.seed(123)
bo <- boot(cdiac, statistic = fit_stat, R = 10, lambda = 0.44)
ci <- t(sapply(seq_len(nrow(cdiac)), function(i)
  boot.ci(bo, conf = 0.95, type = "norm", index = i)$normal[-1]))
band <- data.frame(year = cdiac$year, annual = cdiac$annual,
                   est = bo$t0, lower = ci[, 1], upper = ci[, 2])
```

```{r}
#| label: neariso-plot
#| fig-cap: "Nearly-isotonic fit with a bootstrap 95% band."
ggplot(band, aes(year)) +
  geom_point(aes(y = annual), color = "grey60", size = 0.8) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.3, fill = "steelblue") +
  geom_line(aes(y = est), color = "steelblue", linewidth = 1) +
  labs(x = "Year", y = "Temperature anomaly") +
  theme_minimal()
```

The blue curve is the fit; the band is its bootstrap uncertainty --- inference for an
estimator that has no formula, obtained by *re-solving* the convex problem on
resampled data. (This is exactly the kind of place where the @sec-dpp parameter trick
pays off: every bootstrap replicate is the same problem with new data.)

## Exercise

**Exercise (⭐).** The penalty weight $\lambda = 0.44$ controls smoothness. Refit the
temperature series at $\lambda = 0$ and a large $\lambda = 10$. What do the two
extremes correspond to?

::: {.callout-tip collapse="true"}
## Solution
```{r}
#| label: ex-lambda
fit0  <- neariso_fit(cdiac$annual, 0)
fit10 <- neariso_fit(cdiac$annual, 10)
c(lambda0_matches_data = max(abs(fit0 - cdiac$annual)) < 1e-4,
  lambda10_is_monotone = all(diff(fit10) >= -1e-4))
```
At $\lambda = 0$ there is no penalty, so the fit just reproduces the data. As
$\lambda \to \infty$ the penalty forces the downward steps to zero --- a fully
**isotonic** (monotone non-decreasing) fit.
:::

## Takeaways

- A convex model (**total variation**) reconstructs a damaged image with no
  machine-learning training --- just "match the known pixels, stay smooth."
- The **bootstrap** gives confidence bands for *any* CVXR estimator by re-solving on
  resampled data --- no formula required.
- Both are everyday convex optimization; you now have the vocabulary to build them.

See [TV In-Painting](https://cvxr.rbind.io/examples/applications/tv-inpainting.html)
and [Near-Isotonic Regression](https://cvxr.rbind.io/examples/regression/near-isotonic-and-near-convex-regression.html).
