# The Bootstrap: Inference for Optimization-Based Estimators {#sec-bootstrap}

<!-- LIVE. Source: near-isotonic-and-near-convex-regression.qmd (cdiac ships with
     CVXR). Split out of the old combined showcase chapter and expanded, because the
     bootstrap is a general-purpose inference tool for *any* CVXR estimator. -->

::: callout-note
## What you'll be able to do
Attach **standard errors and confidence bands** to *any* CVXR estimate --- even one
with no closed-form formula --- by re-solving the problem on resampled data. This is
the bridge from optimization back to **statistical inference**.
:::

```{r}
#| label: setup
#| include: false
library(CVXR)
library(ggplot2)
library(boot)
```

## Why this matters

Almost every model in this book produces a **point estimate**: solve the problem, read
off `value(beta)`, done. But a point estimate without a sense of its **uncertainty** is
only half of a statistical answer. For ordinary least squares there is a textbook
standard-error formula. For a lasso, a quantile fit, an isotonic fit, or any of the
custom objectives CVXR makes easy, there usually *isn't* one --- the estimator is
defined implicitly, as "the solution of this optimization problem."

The **bootstrap** sidesteps the missing formula entirely. Its premise is simple: the
sampling variability of an estimator can be approximated by **refitting it on resampled
data**. Because a CVXR estimate is just a deterministic function of the data --- *run
the solver* --- it slots straight into the bootstrap with nothing special required.
That makes the bootstrap a near-universal companion to convex modeling:

> **If you can solve it once, you can bootstrap it.**

## A fit with no standard-error formula

Our estimator is a **nearly-isotonic** fit: least squares plus a penalty on *downward*
steps, so it prefers a mostly-increasing curve without forcing strict monotonicity:

$$\underset{\beta}{\mbox{minimize}}\quad \tfrac12\sum_i (y_i-\beta_i)^2
   + \lambda \sum_i (\beta_i-\beta_{i+1})_+ .$$

We apply it to global-mean temperature anomalies (`cdiac`, shipped with CVXR). The fit
is a one-liner; crucially, it is also a plain R function of a data vector --- exactly
the shape the bootstrap needs:

```{r}
#| label: neariso-fn
data(cdiac)
neariso_fit <- function(y, lambda) {
  beta <- Variable(length(y))
  prob <- Problem(Minimize(0.5 * sum((y - beta)^2) + lambda * sum(pos(-diff(beta)))))
  psolve(prob)
  value(beta)
}
```

## The bootstrap recipe

The `boot` package does the resampling bookkeeping; we just supply a statistic that
**refits on each resample**. Every replicate is a fresh solve, so we keep the number of
replicates modest:

```{r}
#| label: neariso-boot
fit_stat <- function(data, idx, lambda) {
  s <- data[idx, ]
  s <- s[order(s$year), ]          # bootstrap resamples rows; restore time order
  neariso_fit(s$annual, lambda)
}

set.seed(123)
bo <- boot(cdiac, statistic = fit_stat, R = 100, lambda = 0.44)
```

`bo$t0` is the fit on the original data; the `R = 100` rows of `bo$t` are the refits on
resampled data. From their spread we read off a 95% confidence band, one point at a
time:

```{r}
#| label: neariso-band
ci <- t(sapply(seq_len(nrow(cdiac)), function(i)
  boot.ci(bo, conf = 0.95, type = "norm", index = i)$normal[-1]))
band <- data.frame(year = cdiac$year, annual = cdiac$annual,
                   est = bo$t0, lower = ci[, 1], upper = ci[, 2])
```

```{r}
#| label: neariso-plot
#| fig-cap: "Nearly-isotonic fit (blue) with a bootstrap 95% confidence band."
ggplot(band, aes(year)) +
  geom_point(aes(y = annual), color = "grey60", size = 0.8) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.3, fill = "steelblue") +
  geom_line(aes(y = est), color = "steelblue", linewidth = 1) +
  labs(x = "Year", y = "Temperature anomaly") +
  theme_minimal()
```

The blue curve is the fit; the band is its bootstrap uncertainty --- genuine inference
for an estimator that has **no formula**, obtained purely by re-solving the convex
problem on resampled data. The band widens where the data are sparse or noisy and
tightens where they pin the curve down, exactly as inference should.

## Bootstrapping is a parameter sweep

Every replicate solves *the same problem* with *new data* --- precisely the situation
the `Parameter` trick of @sec-dpp was built for. Promote the data (and $\lambda$) to
`Parameter`s, build the problem once, and each refit becomes a cheap re-solve instead
of a fresh compilation. With hundreds or thousands of replicates that saving compounds,
turning the bootstrap from "too slow to bother" into routine. The bootstrap is one of
the best reasons to learn DPP.

## Exercise

**Exercise (⭐).** The penalty weight $\lambda = 0.44$ controls smoothness. Refit the
temperature series at $\lambda = 0$ and a large $\lambda = 10$. What do the two
extremes correspond to?

::: {.callout-tip collapse="true"}
## Solution
```{r}
#| label: ex-lambda
fit0  <- neariso_fit(cdiac$annual, 0)
fit10 <- neariso_fit(cdiac$annual, 10)
c(lambda0_matches_data = max(abs(fit0 - cdiac$annual)) < 1e-4,
  lambda10_is_monotone = all(diff(fit10) >= -1e-4))
```
At $\lambda = 0$ there is no penalty, so the fit just reproduces the data. As
$\lambda \to \infty$ the penalty forces the downward steps to zero --- a fully
**isotonic** (monotone non-decreasing) fit.
:::

## Takeaways

- The **bootstrap** gives standard errors and confidence bands for *any* CVXR
  estimator by re-solving on resampled data --- no formula required.
- The recipe is mechanical: wrap your `psolve()` fit in a function of the data, hand it
  to `boot`, and summarize the replicates.
- Because every replicate is the **same problem with new data**, the @sec-dpp
  `Parameter` trick makes a large bootstrap genuinely cheap.

See [Near-Isotonic Regression](https://cvxr.rbind.io/examples/regression/near-isotonic-and-near-convex-regression.html)
for the underlying model.
