# Beyond Convex: Nonlinear Programming {#sec-dnlp}

<!-- HANDS-ON. CRAN: needs Uno + sparsediff (both in the setup install line;
     ipopt dropped). Source: cvxr_docs/examples/dnlp/tutorial.qmd, trimmed.
     Feature-gated as a graceful fallback if a participant lacks the packages. -->

::: callout-note
## What you'll be able to do
Recognize a **smooth but nonconvex** problem, send it down CVXR's nonlinear path
with `psolve(nlp = TRUE)`, and understand why a nonlinear solve gives only a *local*
guarantee --- and what to do about it.
:::

::: callout-tip
## You already have what you need
This chapter uses two CRAN packages beyond CVXR's built-in solvers --- `sparsediff`
(automatic derivatives) and `Uno` (the nonlinear solver) --- and you installed both
in the [setup](index.qmd#installation). So it is **hands-on, like every other
chapter**. (The `ipopt` backend is optional and not on CRAN; you do not need it.)
:::

```{r}
#| label: feature-gate
#| include: false
library(CVXR)
run_ok <- utils::packageVersion("CVXR") >= "1.9.1" &&
          requireNamespace("sparsediff", quietly = TRUE) &&
          requireNamespace("Uno", quietly = TRUE)
knitr::opts_chunk$set(eval = run_ok)
```

## From convex to merely smooth

Everything so far stayed inside the convex world, where a solution is *guaranteed*
global. **Disciplined Nonlinear Programming (DNLP)**, new in CVXR 1.9.1, relaxes
that: it handles **smooth** problems built from differentiable atoms, even when they
are not convex. You build the problem the usual way, check it with `is_dnlp()`, and
solve with `psolve(prob, nlp = TRUE)`. Every DCP problem is also a DNLP, so the
convex world sits inside this larger one.

```{r}
#| label: available
c(sparsediff = requireNamespace("sparsediff", quietly = TRUE),
  Uno        = requireNamespace("Uno", quietly = TRUE))
```

A convex problem sent down the nonlinear path just works:

```{r}
#| label: first
x <- Variable(2)
prob <- Problem(Minimize(sum_squares(x - c(1, 2))))
is_dnlp(prob)
psolve(prob, nlp = TRUE, solver = "UNO", logger = "SILENT")
value(x)
```

## A genuinely nonconvex problem

Now something DCP *cannot* express. Maximizing a convex quadratic is nonconvex ---
but it is smooth, hence a DNLP. A classic instance is maximizing $x^\top A x$ over
the unit sphere, whose optimum is the **largest eigenvalue of $A$** --- giving us a
known answer to check against:

$$\text{maximize}\ \ x^\top A x \qquad \text{subject to}\qquad \|x\|_2^2 = 1.$$

```{r}
#| label: eigen
set.seed(0)
n <- 3
A <- matrix(rnorm(n * n), n); A <- t(A) %*% A   # random PSD matrix

x <- Variable(n)
prob <- Problem(Maximize(quad_form(x, A)), list(sum_squares(x) == 1))
c(is_dcp = is_dcp(prob), is_dnlp = is_dnlp(prob))   # not DCP, but DNLP

value(x) <- rep(1, n)                                # a starting point
opt <- psolve(prob, nlp = TRUE, solver = "UNO", logger = "SILENT")
c(dnlp = opt, max_eigenvalue = max(eigen(A, only.values = TRUE)$values))
```

The nonlinear solver lands exactly on the largest eigenvalue --- a nonconvex problem
solved from R, in the same modeling language.

## The catch: only a *local* guarantee

This power comes with a caveat you must respect. Unlike convex optimization, a
nonlinear solver returns a **local** optimum that can depend on where it starts ---
there is no global guarantee. Consider $\sin(x) + \tfrac{1}{10}x^2$ on $[-10, 10]$,
which has several local minima:

```{r}
#| label: bestof
set.seed(1)
x <- Variable(bounds = list(-10, 10))
prob <- Problem(Minimize(sin(x) + 0.1 * x^2))

best <- psolve(prob, nlp = TRUE, solver = "UNO", logger = "SILENT", best_of = 5)
c(best_objective = best, x = value(x))
```

`best_of = 5` solves from five random starting points and keeps the best --- a simple,
effective hedge against local minima. The lesson: **convex problems give you the
global optimum for free; nonconvex ones make you work for it** (good starting points,
random restarts, and a healthy skepticism of any single solve).

::: callout-warning
On a nonconvex problem an NLP solver may return a local (not global) optimum, give a
result that depends on the starting point, or fail to find a feasible point. Supply a
starting point with `value(x) <- ...`, or use `best_of` to try several.
:::

## Smooth atoms DCP doesn't have

DNLP adds classically differentiable atoms that are not part of the convex grammar
--- `sin`, `cos`, `tan`, `sinh`, `tanh`, `asinh`, `atanh`, `normcdf`, `prod` --- and
lets existing smooth atoms (`exp`, `log`, `quad_form`, `quad_over_lin`, …) be used in
either curvature direction. Three predicates classify an expression under the smooth
grammar:

```{r}
#| label: predicates
x <- Variable()
classify <- function(e) c(smooth      = is_smooth(e),
                          lin_convex  = is_linearizable_convex(e),
                          lin_concave = is_linearizable_concave(e))
rbind(`sin(x)`  = classify(sin(x)),
      `abs(x)`  = classify(abs(x)),
      `-abs(x)` = classify(-abs(x)))
```

A smooth atom like `sin(x)` linearizes in *both* directions, so it may appear
anywhere; nonsmooth convex/concave atoms keep the curvature DCP assigns them.

## Exercises

**Exercise 1.** See the local-optimum trap for yourself. Solve
$\sin(x) + \tfrac{1}{10}x^2$ from a *single* starting point at `x0 = 4`, then
compare with `best_of = 20`. Do they agree?

::: {.callout-tip collapse="true"}
## Solution
```{r}
#| label: ex-local
x <- Variable(bounds = list(-10, 10))
prob <- Problem(Minimize(sin(x) + 0.1 * x^2))

value(x) <- 4                                              # a poor starting point
one <- psolve(prob, nlp = TRUE, solver = "UNO", logger = "SILENT")

value(x) <- NULL                    # clear the fixed start so best_of can randomize
set.seed(2)
many <- psolve(prob, nlp = TRUE, solver = "UNO", logger = "SILENT", best_of = 20)
c(from_x0_4 = one, best_of_20 = many)
```
The single solve from `x0 = 4` gets **stuck in a nearby local minimum** (objective
$\approx 0.83$); the random restarts escape it and find the global one
($\approx -0.79$). This is the whole story of nonconvex optimization in one example.
:::

**Exercise 2 (⭐).** Turn on `verbose = TRUE` for one of the solves above and read
UNO's log. How is it different from the convex-solver output you saw in
@sec-gentle-intro?

::: {.callout-tip collapse="true"}
## Solution
```{r}
#| label: ex-verbose
#| eval: false
psolve(prob, nlp = TRUE, solver = "UNO", verbose = TRUE)
```
You will see UNO *reformulate* the model and run a nonlinear iteration (a Newton or
SQP method), rather than the cone solver's interior-point iterations. Same
`verbose = TRUE` switch, a different engine underneath.
:::

## Takeaways

- **DNLP** extends CVXR beyond convexity to **smooth** problems via
  `psolve(nlp = TRUE)`; `is_dnlp()` checks membership, and every DCP problem
  qualifies.
- It unlocks genuinely nonconvex models (and smooth atoms like `sin`/`cos`) --- but
  trades the **global guarantee** for a local one; use starting points and `best_of`.
- It needs the CRAN packages `sparsediff` + `Uno` (the practical R default; `ipopt`
  is optional and not on CRAN).

For the full treatment --- duals on the NLP path, random-restart details, the smooth
atom table --- see the [DNLP Tutorial](https://cvxr.rbind.io/examples/dnlp/tutorial.html).
