# DGP and DQCP {#sec-dgp-dqcp}

<!-- MENTION ONLY (pointer chapter), one tiny verified example each. Source:
     cvxr_docs/examples/dgp/*, dqcp/*. Kept deliberately brief. -->

::: callout-note
## What you'll learn about
Two more "disciplined" dialects beyond DCP --- enough to know they exist and when to
reach for them. Both are covered in depth on the CVXR website.
:::

CVXR's convex grammar (DCP) is not the only disciplined rule set. Two siblings extend
the same idea --- *compose from a library, get a guarantee* --- to other problem
classes. We mention them briefly here; follow the links for the full treatment.

## Disciplined Geometric Programming (DGP)

Some problems are not convex in the variables but become convex after a
**log-log** change of variables --- *geometric programs*, built from positive
variables, monomials, and posynomials. They appear in engineering design, and in
statistics in problems involving products and ratios of positive quantities. You
write them naturally and pass `gp = TRUE`. The classic example maximizes the volume
of a box subject to area limits:

```{r}
#| label: setup
#| include: false
library(CVXR)
library(ggplot2)
library(survival)
source("_check_status.R")
theme_set(theme_bw())
```

```{r}
#| label: dgp
x <- Variable(pos = TRUE); y <- Variable(pos = TRUE); z <- Variable(pos = TRUE)
prob <- Problem(Maximize(x * y * z),
                list(2 * (x * z + y * z) <= 10, x * y <= 4))
psolve(prob, gp = TRUE)
round(c(x = value(x), y = value(y), z = value(z)), 3)
```

The objective `x * y * z` multiplies three variables --- not DCP at all --- yet DGP
solves it to global optimality. See the
[DGP Tutorial](https://cvxr.rbind.io/examples/dgp/tutorial.html).

### A worked example: ordering survival curves

DGP is not only for engineering design; it appears in statistics wherever a model is
built from products and ratios of positive quantities. Here is one such problem, from
@LimKimWang:2009.

Suppose we estimate survival curves $S_1, \ldots, S_N$ for several populations and
know on scientific grounds that they are **stochastically ordered**: population $a$
dies earlier than $b$, meaning $S_a(t) \le S_b(t)$ at every $t$ (written $S_a \preceq
S_b$). The unconstrained nonparametric estimates --- the Kaplan--Meier curves --- often
*cross* where they should not. @LimKimWang:2009 observed that imposing the ordering
turns the estimation into a geometric program, which `CVXR` solves directly.

Take **right-censored** data. For population $i$, at each distinct failure time
$t_{ij}$ let $n_{ij}$ subjects be at risk and $d_{ij}$ of them fail. Write the one-step
conditional survival probability $p_{ij} = S_i(t_{ij}) / S_i(t_{i,j-1})$ and $q_{ij} =
1 - p_{ij}$, so survival is the running product $S_i(t_{ij}) = \prod_{r \le j} p_{ir}$
(the Kaplan--Meier form). The likelihood is the **monomial**
$$
\mathcal L = \prod_{i} \prod_{j} p_{ij}^{\,n_{ij} - d_{ij}}\, q_{ij}^{\,d_{ij}},
$$
which is log-log affine and so drops straight into a GP. Three ingredients complete the
model:

- **Sum-to-one**, relaxed from $p_{ij} + q_{ij} = 1$ to the posynomial inequality
  $p_{ij} + q_{ij} \le 1$ (a GP admits only monomial equalities). It is tight at the
  optimum, since both exponents are nonnegative.
- **Monotonicity** of each survival curve, which is *automatic* here: $S_i$ is a
  product of factors $p \le 1$, so it is nonincreasing with no extra constraint.
- **Ordering** $S_a \preceq S_b$: at each pooled failure time $t$, the monomial
  $\bigl(\prod_{t_{ar} \le t} p_{ar}\bigr)\bigl(\prod_{t_{br} \le t} p_{br}\bigr)^{-1}
  \le 1$.

Everything is a monomial except the sum-to-one posynomial, so `is_dgp()` holds. We use
the oropharyngeal-carcinoma trial of Kalbfleisch and Prentice, grouping patients by
tumor stage `T` and nodal stage `N`:

```{r}
#| label: surv-data
oropharynx <- readRDS("data/oropharynx.rds")

grp_counts <- function(T, N) {                          # discrete-hazard counts
  s  <- subset(oropharynx, tstage == T & nstage == N)
  dt <- sort(unique(s$days[s$status == 1]))             # distinct failure times
  data.frame(t = dt,
             d = sapply(dt, function(x) sum(s$days == x & s$status == 1)),  # failures
             n = sapply(dt, function(x) sum(s$days >= x)))                  # at risk
}
```

`fit_rc()` assembles the monomial likelihood, the relaxed sum-to-one constraints, and
one ordering monomial per pooled failure time for each requested pair `c(a, b)`
(meaning $S_a \preceq S_b$):

```{r}
#| label: surv-fit
fit_rc <- function(groups, order = list()) {
  G <- length(groups)
  p <- lapply(groups, function(g) Variable(nrow(g), pos = TRUE))
  q <- lapply(groups, function(g) Variable(nrow(g), pos = TRUE))
  mono <- function(v, a) Reduce(`*`, lapply(which(a != 0), function(j) v[j]^a[j]))
  Scum <- function(i, ix) if (length(ix)) prod_entries(p[[i]][ix]) else 1  # empty prod = 1

  objective <- Reduce(`*`, lapply(seq_len(G), function(i)
                 mono(p[[i]], groups[[i]]$n - groups[[i]]$d) * mono(q[[i]], groups[[i]]$d)))
  constraints <- lapply(seq_len(G), function(i) p[[i]] + q[[i]] <= 1)

  for (o in order) {                    # S_a(t) <= S_b(t) at every pooled failure time
    a <- o[1]; b <- o[2]
    for (tk in sort(unique(c(groups[[a]]$t, groups[[b]]$t)))) {
      ia <- which(groups[[a]]$t <= tk); ib <- which(groups[[b]]$t <= tk)
      constraints <- c(constraints, Scum(a, ia) * Scum(b, ib)^(-1) <= 1)
    }
  }
  prob <- Problem(Maximize(objective), constraints)
  val  <- psolve(prob, gp = TRUE)
  check_solver_status(prob)
  list(is_dgp = is_dgp(prob), logLik = log(val),
       S = lapply(seq_len(G), function(i) cumprod(value(p[[i]]))))  # S = running product
}
```

With no ordering the GP must return the ordinary Kaplan--Meier estimator --- a
reassuring check:

```{r}
#| label: surv-km-check
g   <- grp_counts(3, 3)
fit <- fit_rc(list(g))
km  <- survfit(Surv(days, status) ~ 1,
               data = subset(oropharynx, tstage == 3 & nstage == 3))
max(abs(fit$S[[1]] - summary(km, times = g$t)$surv))    # ~ 0: GP == Kaplan-Meier
```

Now the ordering. Take the three T = 3 groups differing in nodal involvement,
$(T,N) = (3,1), (3,2), (3,3)$, and impose $S_3 \preceq S_2 \preceq S_1$ (more nodal
involvement, worse survival):

```{r}
#| label: surv-fig
#| fig-width: 8
#| fig-height: 3.3
g1 <- list("(3,1)" = grp_counts(3, 1),
           "(3,2)" = grp_counts(3, 2),
           "(3,3)" = grp_counts(3, 3))

rc_free  <- fit_rc(g1)
rc_order <- fit_rc(g1, list(c(3, 2), c(2, 1)))          # S3 <= S2 <= S1

step_rc <- function(fit, groups, panel)
  do.call(rbind, lapply(seq_along(groups), function(i) {
    g <- groups[[i]]; S <- fit$S[[i]]
    data.frame(t = c(0, g$t), S = c(1, S),
               grp = paste0("(T,N)=", names(groups)[i]), panel = panel)
  }))

df <- rbind(step_rc(rc_free,  g1, "(a) Kaplan-Meier (unconstrained)"),
            step_rc(rc_order, g1, "(b) ordered: S3 <= S2 <= S1"))

ggplot(df, aes(t, S, colour = grp)) + geom_step(linewidth = 0.6) +
  facet_wrap(~panel) + coord_cartesian(xlim = c(0, 1850), ylim = c(0, 1)) +
  labs(x = "time (days)", y = "survival probability", colour = NULL) +
  theme(legend.position = "inside", legend.position.inside = c(0.9, 0.8),
        legend.background = element_blank())
```

The unconstrained Kaplan--Meier curves cross --- $(3,2)$ sits above $(3,1)$ for much of
the range, contradicting $S_2 \le S_1$ --- while the ordered fit removes the crossings,
reproducing Figure 1 of @LimKimWang:2009. The ordering costs only a little likelihood:

```{r}
#| label: surv-ll
c(unconstrained = round(rc_free$logLik, 2), ordered = round(rc_order$logLik, 2))
```

The website's
[survival-ordering example](https://cvxr.rbind.io/examples/dgp/survival-ordering.html)
carries this further --- partial (non-total) orders, a stronger **uniform** ordering
that constrains the hazard ratio, and the **interval-censored** case, where the same GP
recipe applies with a different parameterization.

## Disciplined Quasiconvex Programming (DQCP)

A **quasiconvex** function has convex sub-level sets but need not be convex itself
(think of ratios like $\|Ax-b\|/c^\top x$). DCP cannot handle these, but DQCP can,
via a sequence of convex feasibility problems; you pass `qcp = TRUE`. A tiny example
minimizes $(a+1)/\sqrt{a}$, which is quasiconvex on $a > 0$:

```{r}
#| label: dqcp
a <- Variable(pos = TRUE)
prob <- Problem(Minimize((a + 1) / sqrt(a)))
c(is_dqcp = is_dqcp(prob), optimum = psolve(prob, qcp = TRUE), a = value(a))
```

The minimum is $2$ at $a = 1$, found without you having to know the function was
quasiconvex. See the
[DQCP Tutorial](https://cvxr.rbind.io/examples/dqcp/tutorial.html).

## Where this leaves us

You now have the whole map: **DCP** (convex, this book's backbone), **DPP**
(parametrized, for fast re-solves and derivatives), **DGP** (log-log convex),
**DQCP** (quasiconvex), and **DNLP** (smooth nonconvex). Each is the same disciplined
bargain --- follow the grammar, get a solvable problem --- and CVXR checks compliance
for you with the matching `is_*()` predicate. The website's
[examples gallery](https://cvxr.rbind.io) has dozens more worked problems across all
of them.
