# Using Other Solvers

## Goals

- Show how to use non-default solvers like `MOSEK` and `GUROBI`
- Discuss solver peculiarities

## Default Solvers

The default installation of `CVXR` comes with two (imported) open
source solvers:

- [ECOS](https://github.com/embotech/ecos) and its mixed integer
  cousin `ECOS_BB` via the CRAN package
  [ECOSolveR](https://cloud.r-project.org/package=ECOSolveR)
- [SCS](https://github.com/cvxgrp/scs) via the CRAN package
  [scs](https://cloud.r-project.org/package=scs).
  
`CVXR` can also make use of several other open source solvers
implemented in R packages. 

- The linear and mixed integer programming package
  [`lpSolve`](http://lpsolve.sourceforge.net/5.5/) via the 
  [`lpSolveAPI`](https://cloud.r-project.org/package=lpSolveAPI) package
- The linear and mixed integer programming package [`GLPK`](https://www.gnu.org/software/glpk/) via the 
  [`Rglpk`](https://cloud.r-project.org/package=Rglpk) package.
  
Since these are optional, you must install the packages yourself.

```{r}
lapply(list(LPSOLVE = "lpSolveAPI",
            GLPK = "Rglpk"),
       function(x) x %in% installed.packages()[, 1])
```

Once the packages are installed, a call to `installed_solvers` will
display the solvers that `CVXR` has detected.

## Commercial Solvers

A few commercial solvers are also currently supported: [MOSEK](https://www.mosek.com) and
[GUROBI](https://www.gurobi.com).

At the moment, `CVXR` calls vendor Python solver packages via
[`reticulate`](https://cran.r-project.org/package=reticulate), _not R
packages_. Future versions will provide support directly using
[_problem
reductions_](https://web.stanford.edu/~boyd/papers/cvxpy_rewriting.html),
implemented in [`CVXPY` version 1.0](https://www.cvxpy.org/).

Thus, one needs two prerequisites to use these commercial solvers:

- A Python installation in addition to R
- The
  [`reticulate`](https://cran.r-project.org/package=reticulate) R
  package. 
  
We also recommend that you ensure `reticulate` is installed
correctly and working. For example `reticulate::py_eval('1+1')` should return
`2`. 

### Installing `MOSEK`

[MOSEK](https://www.mosek.com) provides an academic version that is
free of charge. As noted in the downloads page, Anaconda users can
install merely via:

```{bash, eval=FALSE}
conda install -c mosek mosek
```

Others can use the `pip` command:

```{bash, eval = FALSE}
pip install -f https://download.mosek.com/stable/wheel/index.html Mosek
```

In addition, the license for the product has to be activated per
instructions on the `Sales` section of the MOSEK web page.

Once activated, you can check that `CVXR` recognizes the solver;
`installed_solvers()` should list `MOSEK`. Otherwise, rinse and repeat
until success.

### Installing `GUROBI`

[GUROBI](https://www.gurobi.com) also provides an academic version
that is free of charge. You must register to receive the license. 

Once registered, install the _Gurobi Optimizer_ software and activate
your license as necessary.

After activation, you can check that `CVXR::installed_solvers()` lists
`GUROBI`. Otherwise, rinse and repeat until success.

### Gotchas

If you have an Anaconda installation in your path, you have
to account for the fact that there may be interactions when using
RStudio and rendering documents. In particular, Anaconda may include
its own version of pandoc and other tools that may conflict with what
Rstudio needs to work properly.

To be concrete, one problem we found was that the `MOSEK` solver was
not recognized as available in this rendered document, even
though the command line interface showed it to be present. Ensuring an
appropriate `PATH` variable solves the problem. 

### Example Session

```{r}
installed_solvers()
```

## Solver Peculiarities

The default solver in `CVXR` is `ECOS`. However, it is not always the
best solver to use. As an example, let us consider again the [catenary
problem](/cvxr_examples/cvxr_catenary/). 

We will change the problem slightly to use a finer discretization from
101 to 501 points.

```{r}
## Problem data
m <- 501
L <- 2
h <- L / (m - 1)

## Form objective
x <- Variable(m)
y <- Variable(m)
objective <- Minimize(sum(y))

## Form constraints
constraints <- list(x[1] == 0, y[1] == 1,
                    x[m] == 1, y[m] == 1,
                    diff(x)^2 + diff(y)^2 <= h^2)

## Solve the catenary problem
prob <- Problem(objective, constraints)
result <- solve(prob)
```

The solution status is no longer optimal.

```{r}
cat("Solution status is", result$status)
```

In such cases, using a different solver may give more accurate
results. Let us try `MOSEK`.

```{r}
if ("MOSEK" %in% installed_solvers()) {
    result <- solve(prob, solver = "MOSEK")
    cat("Solution status is", result$status)
} else {
    cat("Solution not available as MOSEK is not installed!")
}
```

This returns an optimal solution. 

Here again, even commercial solvers differ; `GUROBI`, for example,
does not completely solve the problem and in fact throws an error.


## References
