# The Structure of Atoms

`CVXR` comes with a rich library of atoms that represent common
functions in convex analysis. When combined using the DCP rules, these
atoms are sufficient to model and solve most convex optimization
problems, and we encourage most users to work with the `CVXR`
library. However, it is possible for a sophisticated user to add a new
atom. To do this requires an understanding of mathematical programming
and the S4 class system. In this guide, we go over the basics of
implementing an atom, the quadratic-over-linear function, which should
provide a starting point for potential developers.

## Definition

For any $x \in \mathbf{R}^n$ and $y \in \mathbf{R}$, the
quadratic-over-linear (QoL) function is

$$
  f(x,y) := \frac{\|x\|_2^2}{y} \quad \mbox{with} \quad \mathbf{dom}\;f = \{(x,y) \in \mathbf{R}^n \times \mathbf{R}: y > 0\}.
$$

In `CVXR`, atoms like this (along with variables, constraints, etc) are
represented by S4 class objects. S4 allows us to overload standard
mathematical operations so `CVXR` combines seamlessly with native R
script and other packages. The class for the QoL function is
`QuadOverLin`, defined as

```{r, eval=FALSE}
setClass("QuadOverLin", representation(x = "ConstValORExpr", y = "ConstValORExpr"), contains = "Atom")
```
We also provide a constructor
```{r, eval=FALSE}
quad_over_lin <- function(x, y) { new("QuadOverLin", x = x, y = y) }
```

The `QuadOverLin` class inherits from the `Atom` superclass. It takes
as input two arguments, `x` and `y`, which must be R numeric constants
or `CVXR` `Expression` objects. These input types are encapsulated in
the "ConstaValORExpr" class union. Since `Expression`s themselves may
contain other atoms, this allows us to use the QoL function in nested
compositions.

During initialization, `x` and `y` must be passed to the `Atom`
superclass for processing and validation. Consequently, we override
the default `QuadOverLin` initialization method with

```{r, eval=FALSE}
setMethod("initialize", "QuadOverLin", function(.Object, ..., x, y) {
  .Object@x <- x
  .Object@y <- y
  callNextMethod(.Object, ..., args = list(.Object@x, .Object@y))
})
```

The first two lines save `x` and `y` to their respective slots in
`QuadOverLin`, while the third calls `Atom`'s `initialize` with
`args` equal to the list of arguments. For explanatory purposes,
we reproduce this method below.

```{r, eval=FALSE}
setMethod("initialize", "Atom", function(.Object, ..., args = list(), .size = NA_real_) {
  .Object@args <- lapply(args, as.Constant)
  validate_args(.Object)
  .Object@.size <- size_from_args(.Object)
  callNextMethod(.Object, ...)
})
```

It is not important to understand the details of the above code. Only
recognize that the atom arguments are converted into `CVXR` objects
and saved in a list in the `args` slot. We will be accessing this slot
later.

## Atom Properties

Now that we have created the atom's S4 class, it is time to define
methods that characterize its mathematical properties.

### Mathematical Basics

First, we provide a method for validating the atom's inputs. This
method is called in `initialize` to check whether the arguments make
sense for the atom. For our function $f$, the second argument must be
a scalar, so our validation method looks like

```{r, eval=FALSE}
setMethod("validate_args", "QuadOverLin", function(object) {
  if(!is_scalar(object@args[[2]]))
    stop("[QuadOverLin: validation] y must be a scalar")
})
```

The input `object` is a `QuadOverLin` atom. We access its arguments
via its `args` slot inherited from `Atom`. Given our ordering in
`args`, we know `object@args[[1]]` contains the argument for $x$
and `object@args[[2]]` the one for $y$. Hence, we invoke `is_scalar`,
an `Expression` class method, on the latter to check if $y$ is indeed
scalar. If this check passes, then a call is made up the stack to
`validate_args` in `Atom`.

Notice that we did not check whether $y > 0$. This is because we
generally do not know the value of an argument at the time of
construction. For instance, `y` may contain a `Variable` that we are
solving for in a problem, and we cannot define the problem if
validation fails. It is the user's responsibility to include domain
constraints during problem construction.

To facilitate this, we define a method for that that returns a list 
of `Constraint` objects delineating an atom's domain. For the QoL
function, we need only constrain its second argument to be
positive. Strict inequalities are not supported in `CVXR`, so we
impose a weak inequality $y \geq 0$ as shown below.

```{r, eval=FALSE}
setMethod(".domain", "QuadOverLin", function(object) { list(object@args[[2]] >= 0) })
```

Both validation and domain methods may be dropped if an atom's
arguments span the reals.

The `to_numeric` method is always required. It takes as input an atom and a list `values` containing the numeric values of its arguments (given in the same order as in `args`), then evaluates the atom at these values. For the QoL function, this method is simply

```{r, eval=FALSE}
setMethod("to_numeric", "QuadOverLin", function(object, values) { sum(values[[1]]^2) / values[[2]] })
```

Note the difference between `object@args` and `values`. The former is
a list of `Expression`s, which represent a composition of constants,
variables, and atoms. The latter is a list of R numeric constants like
`numeric` and `matrix`, denoting the values of the corresponding
expressions.

### Dimension, Sign, and Curvature

For DCP analysis to work, we must explicitly define each atom's dimension, 
sign, and curvature. `CVXR` uses this information when applying the 
composition rules to an expression. It is easy to see that $f$ is 
scalar-valued, nonnegative, and convex over its domain. These properties 
are encoded in

```{r, eval=FALSE}
setMethod("size_from_args", "QuadOverLin", function(object) { c(1,1) })
setMethod("sign_from_args",  "QuadOverLin", function(object) { c(TRUE, FALSE) })
setMethod("is_atom_convex", "QuadOverLin", function(object) { TRUE })
setMethod("is_atom_concave", "QuadOverLin", function(object) { FALSE })
```

Here, the `sign_from_args` function returns a vector of two logical
values - the first indicates if the atom's value is positive, and the
second if it is negative. 

We also define whether the atom is weakly
increasing or decreasing in each of its arguments. By taking
derivatives, we find that $f$ is weakly decreasing in $y$, weakly
increasing in $x$ on $\mathbf{R}_+^n$, and weakly decreasing in $x$ on
$\mathbf{R}_-^n$. These properties are spelled out in

```{r, eval=FALSE}
setMethod("is_incr", "QuadOverLin", function(object, idx) { (idx == 1) && is_positive(object@args[[idx]]) })
setMethod("is_decr", "QuadOverLin", function(object, idx) { ((idx == 1) && is_negative(object@args[[idx]])) || (idx == 2) })
```

where `idx` is the index of the argument of interest (`idx = 1` for `x` and `idx = 2` for `y`). 
We call the `Expression` class method `is_positive` (resp. `is_negative`) to determine
if an argument is nonnegative (resp. nonpositive). The user may be
tempted to write `object@args[[idx]] >= 0`, but this is **incorrect**
because it returns a `Constraint` rather than a logical value.

### Additional Characteristics

The methods we have just described are the bare minimum required to
define an atom. Additional methods may be implemented that provide further
functional characterization. For instance, we furnish a method for
computing the gradient of the QoL function,

$$
  \nabla_x f(x,y) = \frac{2x}{y}, \quad \nabla_y f(x,y) = -\frac{\|x\|_2^2}{y^2},
$$

at a point $(x,y)$.

```{r, eval=FALSE}
setMethod(".grad", "QuadOverLin", function(object, values) {
  X <- values[[1]]
  y <- as.numeric(values[[2]])
  if(y <= 0)
    return(list(NA_real_, NA_real_))
  else {
    # DX = 2X/y, Dy = -||X||^2_2/y^2
    Dy <- -sum(X^2)/y^2
    Dy <- Matrix(Dy, sparse = TRUE)
    DX <- 2.0*X/y
    DX <- Matrix(as.numeric(t(DX)), sparse = TRUE)
    return(list(DX, Dy))
  }
})
```

This method calculates the vectors $\nabla_x f(x,y)$ and $\nabla_y
f(x,y))$, wraps each in a sparse `Matrix`, and returns them in a list
ordered exactly as in the input values. If $(x,y) \notin
\mathbf{dom}\;f$, all gradients are set to `NA`.

## Canonicalization

Once `CVXR` verifies a problem is DCP, it converts that problem into a
solver-compatible form. This *canonicalization* process is carried out
through a series of calls to individual atom canonicalizers. 
To implement an atom, we must explicitly derive its canonicalizer for a conic program. 
Below we provide a derivation based on the graph implementation.

A function $g: \mathbf{R}^n \rightarrow \mathbf{R}$ is convex if and
only if its epigraph

$$
  \mathbf{epi}\;g = \{(x,t) \in \mathbf{R}^n \times \mathbf{R}: g(x) \leq t\}
$$
is a convex set. Then, it can be written as
$$
  g(x) = \inf \{t \in \mathbf{R}: (x,t) \in \mathbf{epi}\;g \}.
$$

A similar relationship holds between concave functions and their
hypographs. The *graph implementation* of a function is a
representation of its epigraph or hypograph as a disciplined convex
feasibility problem. This offers an elegant means of defining a
nondifferentiable function in terms of a canonical optimization
problem, which can be directly evaluated by a solver.

For instance, the QoL function can be written as a second-order cone
program (SOCP). Given $(x,y) \in \mathbf{dom}\;f$, the inequality
$f(x,y) \leq t$ is equivalent to

$$
\begin{align*}
  4\|x\|_2^2 &\leq 4ty = (y+t)^2 - (y-t)^2 \\
  (y-t)^2 + \|2x\|_2^2 &\leq (y+t)^2 \\
  \left\|\begin{pmatrix} y-t \\ 2x \end{pmatrix}\right\|_2 &\leq y+t
\end{align*}
$$
so that
$$
  f(x,y) = \inf \left\{t \in \mathbf{R}: \left(\begin{pmatrix} y-t \\ 2x \end{pmatrix}, y+t\right) \in \mathcal{K} \right\},
$$
where $\mathcal{K} := \{(u,v) \in \mathbf{R}^{n+1} \times \mathbf{R}:
\|u\|_2 \leq v\}$ is a second-order cone. Thus, $f(x,y)$ is the
solution to an SOCP, which may be evaluated using any conic
solver. The canonicalizer for the QoL function takes as input $(x,y)$
and outputs the above SOCP.

In `CVXR`, this canonicalizer function is defined as

```{r, eval=FALSE}
QuadOverLin.graph_implementation <- function(arg_objs, size, data = NA_real_) {
  x <- arg_objs[[1]]
  y <- arg_objs[[2]]   # Known to be a scalar.
  t <- create_var(c(1,1))
  two <- create_const(2, c(1,1))
  constraints <- list(SOC(lo.sum_expr(list(y, t)),
                          list(lo.sub_expr(y, t),
                               lo.mul_expr(two, x, x$size))),
                      create_geq(y))
  list(t, constraints)
}
```

It takes as input a list of arguments `arg_objs`, which specify the input values $(x,y)$, the `size` of the resulting expression, and a list of additional `data` required by the atom. 
The first two lines of the function extract $x$ and $y$ from `args`, the third line constructs the `Variable` $t \in \mathbf{R}$, and the fourth line defines the constant $2$. 
The fifth line forms the constraint $\left(\begin{pmatrix} y-t \\ 2x \end{pmatrix}, y+t\right) \in \mathcal{K}$ with a call to the constructor `SOC`. 
If $u \in \mathbf{R}^n$, then `SOC(t,u)` enforces $\|u\|_2 \leq t$. Finally, the canonicalizer returns a list containing
$t$, the epigraph variable, and a list of `Constraint`s from the graph implementation - in this case, the second-order cone constraint and $y \geq 0$.

## Putting It All Together

With this machinery in place, we are ready to use the `QuadOverLin`
atom in a problem. Suppose we are given $A \in \mathbf{R}^{m \times
n}$ and $b \in \mathbf{R}^m$, and we would like to solve

$$
  \begin{array}{ll}
    \mbox{minimize} & f(x,y) \\
    \mbox{subject to} & Ax = b
  \end{array}
$$

with respect to $x \in \mathbf{R}^n$ and $y \in \mathbf{R}_{++}$. In
`CVXR`, we write

```{r, eval=FALSE}
x <- Variable(n)
y <- Variable()
obj <- quad_over_lin(x,y)
constr <- c(domain(obj), A %*% x == b)
prob <- Problem(Minimize(obj), constr)
solve(prob)
```

Once `solve` is invoked, `CVXR` checks if the problem is DCP with calls
to `is_atom_convex`, `is_incr`, and `is_decr` for each atom in the
expression tree. It then converts the problem as modeled to a form compatible 
with the desired solver.

In our example, this requires a transformation into a
cone program. Substituting in the graph implementation of $f$, our
original problem can be written as

$$
  \begin{array}{ll}
    \mbox{minimize} & t \\
    \mbox{subject to} & Ax = b,  \quad (x,y) \in \mathbf{dom}\;f, \quad \left(\begin{pmatrix} y-t \\ 2x \end{pmatrix}, y+t\right) \in \mathcal{K},
  \end{array}
$$

where $t \in \mathbf{R}$ is the additional epigraph variable. If $b
\in \mathbf{null}\;A$, a solution is trivially $x^{\star} = \vec{0}$ with
any $y^{\star} > 0$. Otherwise, the point $y = 0$ is infeasible, so we can
relax the domain constraint to get

$$
    \begin{array}{ll}
    \mbox{minimize} & t \\
    \mbox{subject to} & Ax = b, \quad y \geq 0, \quad \left(\begin{pmatrix} y-t \\ 2x \end{pmatrix}, y+t\right) \in \mathcal{K}.
  \end{array}
$$

In `CVXR`, this SOCP canonicalization is performed automatically via a
call to `QuadOverLin.graph_implementation`. Then, the relevant matrices 
are formed and passed to the selected conic solver. After the solver 
returns $(x^{\star},y^{\star},t^{\star})$, the transformation is reversed 
to obtain the optimal point $(x^{\star},y^{\star})$ and objective value 
$f(x^{\star},y^{\star})$ for the original problem.

