---
title: "Convex Optimization in R Using CVXR"
subtitle: "useR! 2026 · Warsaw · A Hands-On Tutorial"
author: "Anqi Fu · Balasubramanian Narasimhan · Stephen Boyd"
date: 2026-07-06
format:
  revealjs:
    theme: [default, slides.scss]
    slide-number: c/t
    width: 1280
    height: 720
    code-line-numbers: false
    link-external-newwindow: true
    footer: "[cvxr.rbind.io](https://cvxr.rbind.io)"
execute:
  echo: true
  warning: false
  message: false
---

## Why we're here {.smaller}

This is the **opening** to a hands-on session. The real work lives in the
**book** and the **RStudio project** in front of you --- this deck just sets the
stage, then we drop into code.

::: incremental
- **Estimation is often optimization.** You already do this every day.
- A huge share of statistics and ML is, underneath, **convex optimization**.
- **CVXR** lets you write these problems in *natural mathematical notation* and
  have them solved to **global** optimality --- no custom solver, no matching to a
  canned routine.
- Plan for the next hours: run every chunk yourself, modify it, break it, fix it.
:::

::: aside
Tutorial home & all examples: [cvxr.rbind.io](https://cvxr.rbind.io)
:::

## Optimization is everywhere in statistics

We **maximize** a likelihood, **minimize** a loss, or **penalize** a fit to trade
bias for variance.

The problems in this tutorial all have the form

$$
\begin{array}{lll}
  \mbox{minimize}   & f_0(x) & \\
  \mbox{subject to} & f_i(x) \leq 0, & i = 1,\ldots,m\\
                    & Ax = b, &
\end{array}
$$

with variable $x \in \mathbf{R}^n$: an **objective**, **inequality** constraints,
and **equality** constraints. (To maximize, minimize the negative.)

::: aside
Background & on-ramp: [cvxr.rbind.io/examples/intro](https://cvxr.rbind.io/examples/intro.html)
:::

## What makes it *convex* {.smaller}

:::: {.columns}
::: {.column width="55%"}
The objective and inequality functions are **convex**; the equalities are
**affine**.

A function is convex when every **chord** lies *on or above* the graph:

$$f\big(\theta x + (1-\theta) y\big) \le \theta f(x) + (1-\theta) f(y).$$

Picture a **bowl**. Concave is an upside-down bowl; affine is flat (both).
We minimize convex, maximize concave.
:::
::: {.column width="45%"}
```{r}
#| label: chord
#| echo: false
#| fig-width: 4.2
#| fig-height: 3.2
library(ggplot2)
f <- function(x) 0.5 * x^2
xs <- seq(-2.2, 2.2, length.out = 200)
a <- -1.6; b <- 1.3
ggplot(data.frame(x = xs, y = f(xs)), aes(x, y)) +
  geom_line(color = "steelblue", linewidth = 1.1) +
  geom_segment(aes(x = a, y = f(a), xend = b, yend = f(b)),
               color = "red", linewidth = 0.9) +
  geom_point(data = data.frame(x = c(a, b), y = f(c(a, b))),
             color = "red", size = 2.5) +
  theme_minimal(base_size = 14) + labs(x = NULL, y = NULL)
```
:::
::::

::: aside
The DCP rules CVXR enforces: [cvxr.rbind.io/reference/dcp](https://cvxr.rbind.io/reference/dcp.html)
:::

## Why convexity is such good news

::: incremental
- **Every local optimum is global.** Find a solution and it is *the* solution.
- **The feasible region is convex** --- the search space is well-behaved.
- **Solved reliably and fast** by mature algorithms, even nonsmooth, at scale.
:::

. . .

That is why it pays to **recognize** --- or **reformulate into** --- a convex
problem: you trade the uncertainty of general optimization for a *global
guarantee*.

::: aside
More on the landscape: [cvxr.rbind.io/examples/intro](https://cvxr.rbind.io/examples/intro.html)
:::

## You already know many of these {.smaller}

A small sample of methods that are, under the hood, convex optimization:

:::: {.columns}
::: {.column width="50%"}
- ordinary, nonnegative & constrained **least squares**
- **ridge, lasso, elastic-net** penalized regression
- **logistic regression** and **SVMs**
- **quantile** and **Huber (robust)** regression
:::
::: {.column width="50%"}
- **isotonic** & shape-constrained fitting
- **maximum-entropy** & log-concave density
- **portfolio** optimization & finance
- signal processing, control, and more
:::
::::

New methods join the list every year --- and CVXR lets you build them **without
writing a custom solver**.

::: aside
Worked examples by category: [cvxr.rbind.io](https://cvxr.rbind.io)
:::

## CVXR in one screen {.smaller}

$$
\begin{array}{ll}
  \mbox{minimize}   & x^2 + y^2 \\
  \mbox{subject to} & x \geq 0, \quad 2x + y = 1
\end{array}
$$

```{r}
#| label: hello
library(CVXR)

x <- Variable(1)
y <- Variable(1)

objective   <- Minimize(x^2 + y^2)
constraints <- list(x >= 0, 2 * x + y == 1)
prob <- Problem(objective, constraints)

sol <- psolve(prob)
status(prob)              # did it solve?
c(x = value(x), y = value(y))
```

The R *reads like the math*. Every CVXR problem is built this way.

::: aside
Walk through it line by line: [cvxr.rbind.io/examples/gentle-intro](https://cvxr.rbind.io/examples/gentle-intro.html)
:::

## What about non-convex problems?

Most problems are *not* convex --- harder in general, no global guarantee. Two
responses recur in this tutorial:

::: incremental
- A non-convex problem often has a **convex relaxation** --- an approximation you
  *can* solve that gives a useful bound or solution.
- Some are non-convex but **smooth**; CVXR hands these to a nonlinear solver for a
  good local solution (the DNLP chapter).
:::

::: aside
Disciplined nonlinear programming: [cvxr.rbind.io/examples/dnlp/tutorial](https://cvxr.rbind.io/examples/dnlp/tutorial.html)
:::

## How this session runs {.smaller}

::: incremental
- **It's hands-on.** Open the RStudio project; run each chapter's chunks as we go
  --- click *Run* or `Ctrl/Cmd+Enter`.
- **One rule governs everything:** disciplined convex programming (DCP). We learn
  it early, then it just works.
- **Exercises** every chapter --- a base task and a starred ⭐ "modify-and-rerun"
  one, each with a click-to-reveal solution.
- **CRAN only.** `install.packages(c("CVXR", "scip", "Uno", "sparsediff"))` plus a
  few helpers. Run `setup_check.R` before we start.
- We finish with **Bring-Your-Own-Problem** --- translate something *you* care about.
:::

::: aside
Everything expanded, with more examples: [cvxr.rbind.io](https://cvxr.rbind.io)
:::

## Let's begin

Open the book at **A Gentle Introduction** and run the first chunk.

<br>

**Resources**

- The book & all examples: [cvxr.rbind.io](https://cvxr.rbind.io)
- The package paper: Fu, Narasimhan & Boyd (2020), *JSS* **94**(14)
- The standard reference: Boyd & Vandenberghe, *Convex Optimization* (2004)

<br>

*Convex Optimization in R Using CVXR* · Fu, Narasimhan & Boyd
