---
title: "Disciplined Convex Programming"
subtitle: "useR! 2026 · Warsaw · Chapter 3"
format:
  revealjs:
    theme: [default, ../slides.scss]
    slide-number: c/t
    width: 1280
    height: 720
    scrollable: true
    code-line-numbers: false
    link-external-newwindow: true
    footer: "[cvxr.rbind.io](https://cvxr.rbind.io)"
execute:
  echo: true
  eval: true
  warning: false
  message: false
  freeze: false
bibliography: [../../references.bib, ../../cvxr.bib]
knitr:
  opts_knit:
    root.dir: "/Users/naras/GitHub/cvxr_tutorial"
---


<!-- LIVE (backbone). Source: cvxr_docs/reference/dcp.qmd, reference/visualization.qmd;
     2019 ch.04, ch.06. Pays off the invalid-problem error from Chapter 1. -->

::: callout-note
## What you'll be able to do
Read the **sign** and **curvature** of any CVXR expression, understand the one
composition rule that decides whether CVXR accepts a problem, recognize the two
legal objective forms and three legal constraint forms, **reformulate** a rejected
problem into an equivalent accepted one, and use `visualize()` to locate a
violation.
:::

```{r}
#| label: setup
#| include: false
library(CVXR)
```

## Why was that problem rejected?

At the end of *A Gentle Introduction*, CVXR refused to `Maximize(x^2 + y^2)`. It was not
being fussy: maximizing a bowl-shaped (convex) function has no finite solution in
general, and more importantly CVXR will only accept problems it can *certify* are
convex. The certificate is a small rule set called **Disciplined Convex Programming
(DCP)**. This chapter is the one piece of theory in the book --- everything else
rests on it. The good news: the rules are mechanical, and CVXR checks them for you.

For an interactive companion, see [dcp.stanford.edu](https://dcp.stanford.edu).

## Expressions carry a sign and a curvature

A CVXR **expression** is built from variables, constants, the arithmetic operators
`+ - * %*% /`, and a library of atoms (`abs`, `sqrt`, `square`, `p_norm`, ...).
DCP tags every (sub)expression with a **sign** and a **curvature**, computed from
its parts. Both tags are *always correct*, though the simple analysis sometimes
says "unknown" when a deeper argument could do better.

### Sign

Each expression is flagged **nonnegative**, **nonpositive**, **zero**, or
**unknown**:

```{r}
#| label: sign
x <- Variable()
expr_sign(x)            # a free variable could be anything
expr_sign(abs(x))       # absolute value is nonnegative
expr_sign(Constant(-3)) # constants are read off their value
```

### Curvature

Each expression is **constant**, **affine**, **convex**, **concave**, or
**unknown**. The picture to keep in mind:

- **convex** = upward-curving, **concave** = downward-curving, **affine** = flat.
- constant $\subset$ affine $\subset$ both convex and concave.

```{r}
#| label: curvature
curvature(x)                # a variable is affine
curvature(abs(x))           # convex
curvature(sqrt(x))          # concave
curvature(2 * x + 1)        # affine
```

There are also yes/no predicates, handy in code:

```{r}
#| label: predicates
c(convex  = is_convex(abs(x)),
  concave = is_concave(sqrt(x)),
  affine  = is_affine(2 * x + 1))
```

## The one rule: composition

Every curvature tag comes from a single **composition rule**. To conclude that
$f(e_1,\ldots,e_n)$ is **convex**, $f$ must be convex *and* for each argument $e_i$
one of these holds:

- $f$ is **increasing** in $e_i$ and $e_i$ is convex, or
- $f$ is **decreasing** in $e_i$ and $e_i$ is concave, or
- $e_i$ is affine or constant.

(The mirror image gives **concave**.) Whether $f$ is increasing or decreasing can
depend on sign --- `abs` increases for positive arguments and decreases for
negative ones, which is exactly why sign analysis matters.

Let's watch it work on `sqrt(1 + square(x))`:

```{r}
#| label: compose
curvature(square(x))            # square is convex
curvature(1 + square(x))        # affine(+) of convex -> convex
curvature(sqrt(1 + square(x)))  # sqrt needs a CONCAVE argument -> UNKNOWN
```

`square(x)` is convex, so `1 + square(x)` is convex. But `sqrt` is concave *and
increasing*, so by the rule it can only accept a **concave** argument. Handed a
convex one, DCP cannot certify the result --- so it reports **unknown**.

## When DCP says no but the function is fine: reformulate

Here is the subtle part. `sqrt(1 + square(x))` genuinely *is* a convex function of
`x` --- DCP just can't see it through that particular expression. The fix is to
**rewrite it** as something DCP recognizes. The same quantity is the Euclidean norm
of the vector $[1, x]$:

```{r}
#| label: reformulate
curvature(p_norm(vstack(1, x), 2))   # CONVEX -- now certified
```

Same value, accepted formulation. *Reformulating a rejected expression into an
equivalent accepted one is the core CVXR skill* --- you will do it often, and the
atom [function reference](https://cvxr.rbind.io/reference/functions.html) is your
toolbox.

## DCP problems

A problem is **DCP** --- and therefore convex and solvable --- when its objective
has one of two forms and every constraint has one of three:

| Objective | Constraints |
|---|---|
| `Minimize(convex)` | `affine == affine` |
| `Maximize(concave)` | `convex <= concave` |
| | `concave >= convex` |

Check any problem, objective, or constraint with `is_dcp()`:

```{r}
#| label: is-dcp
y <- Variable()

prob1 <- Problem(Minimize((x - y)^2), list(x + y >= 0))
prob2 <- Problem(Maximize(sqrt(x - y)), list(2 * x - 3 == y, x^2 <= 2))
c(prob1 = is_dcp(prob1), prob2 = is_dcp(prob2))

# A non-DCP objective and a non-DCP constraint:
c(max_sq   = is_dcp(Maximize(x^2)),
  bad_con  = is_dcp(sqrt(x) <= 2))
```

Call `psolve()` on a non-DCP problem and CVXR stops you:

```{r}
#| label: nondcp-solve
#| error: true
psolve(Problem(Minimize(sqrt(x))))
```

## Finding the violation with `visualize()`

For anything bigger than a toy, `visualize()` draws the expression tree with each
node's curvature, flagging the offending node so you know *where* to reformulate:

<!-- STATIC until CVXR >= 1.9.3 is on CRAN: 1.9.2 errors on this call (display
     bug with rational power exponents; fixed for 1.9.3). Mirrors 03-dcp.qmd in
     the book; same restoration recipe. -->

```{r}
#| label: visualize
#| eval: false
visualize(Problem(Minimize(sqrt(1 + square(x)))))
```

::: {.callout-note}
Errors on CVXR 1.9.2 (display bug, fixed in 1.9.3) — output below pre-rendered
with the fixed version.
:::

## `visualize()` output (pre-rendered)

```
── Expression Tree (MINIMIZE) ──────────────────────────────────────────
t_3 = PowerApprox(...)  [unknown, \mathbb{R}_+, 1x1]
\-- t_2 = AddExpression(...)  [convex, \mathbb{R}_+, 1x1]
    |-- 1  [constant, 1x1]
    \-- t_1 = PowerApprox(...)  [convex, \mathbb{R}_+, 1x1]
        \-- var1  [affine, 1x1]
── DCP ANALYSIS ────────────────────────────────────────────────────────
  ✗ Problem is NOT DCP compliant.
  Objective (Minimize): requires convex expression
    ✗ PowerApprox: unknown  [1x1]
      arg 1: convex, increasing  (cvx-rule ✓, ccv-rule ✗)
      => PowerApprox is concave, but DCP concave composition fails at arg 1.
      Hint: var1 must be nonneg (>= 0)
── SMITH FORM ──────────────────────────────────────────────────────────
  t_{1} = phi^{(.)^2}({var1})
  t_{2} = {1} + {t_{1}}
  t_{3} = phi^{(.)^{1/2}}({t_{2}})
── RELAXED SMITH FORM ───────────────────────────────────────────────────
  t_{1} >= phi^{(.)^2}({var1})
  t_{2} = {1} + {t_{1}}
  t_{3} = phi^{(.)^{1/2}}({t_{2}})
── CONIC FORM ───────────────────────────────────────────────────────────
  (\frac{1+t_{1}}{2},  \frac{1-t_{1}}{2},  {var1}) in Q^3
  t_{2} = {1} + {t_{1}}
  t_{3} = phi^{(.)^{1/2}}({t_{2}})    [conic form: see canonicalizer]
```

See the [Visualization page](https://cvxr.rbind.io/reference/visualization.html)
for the full feature set (curvature coloring, fix hints, HTML trees).

## Bonus: bounds propagation

Since CVXR 1.9.1, `get_bounds()` propagates interval bounds through any expression
--- useful for sanity-checking a model:

```{r}
#| label: bounds
xb <- Variable(3, bounds = list(-1, 2))
get_bounds(2 * xb + 1)   # affine map of [-1, 2]  ->  [-1, 5]
get_bounds(abs(xb))      # through an atom        ->  [0, 2]
```

## Exercises

**Exercise 1.** Without running it, predict the curvature of `square(sqrt(x))` for
`x >= 0`, then check with `curvature()`. Why does DCP report what it does?

::: {.callout-tip collapse="true"}
## Solution
```{r}
#| label: ex-compose
curvature(square(sqrt(x)))
```
`sqrt(x)` is concave; `square` is convex and (for nonnegative arguments)
increasing, so it wants a *convex* argument. A concave argument breaks the rule, so
DCP returns **unknown** --- even though the composition equals `x`, which is affine.
Reformulating (here, just writing `x`) would fix it.
:::

**Exercise 2.** Is `Problem(Maximize(-square(x) - square(y)), list(x + y == 1))`
DCP? Predict, then check, then solve it.

::: {.callout-tip collapse="true"}
## Solution
```{r}
#| label: ex-dcp
p <- Problem(Maximize(-square(x) - square(y)), list(x + y == 1))
is_dcp(p)
psolve(p)
c(x = value(x), y = value(y))
```
`square` is convex, so `-square(x)` and `-square(y)` are each concave, and a **sum
of concave expressions is concave** --- so the objective `-square(x) - square(y)` is
concave. The constraint is `affine == affine`, so the problem is `Maximize(concave)`
subject to a legal constraint --- DCP, with optimum $x = y = 1/2$.
:::

**Exercise 3 (⭐).** The expression `x^2 / y` (with `y > 0`) is convex --- the
quadratic-over-linear function. DCP cannot see it from `x^2 / y` (division by a
non-constant is rejected). Find a CVXR atom that expresses it. *Hint:* search the
function reference for "quad over lin".

::: {.callout-tip collapse="true"}
## Solution
The atom is `quad_over_lin(x, y)` $= x^2/y$, certified convex:
```{r}
#| label: ex-qol
yp <- Variable(pos = TRUE)
curvature(quad_over_lin(x, yp))
```
This is the reformulation skill again: the *function* is convex; you just need the
formulation DCP recognizes.
:::

## Takeaways

- Every expression has a **sign** and a **curvature**, computed by one
  **composition rule**.
- A problem is solvable when it is **`Minimize(convex)`** or **`Maximize(concave)`**
  with constraints `affine == affine`, `convex <= concave`, or `concave >= convex`.
- `is_dcp()` checks it; `visualize()` shows *where* a violation is.
- When a convex function is rejected, **reformulate** it into an equivalent atom ---
  the single most useful habit in CVXR.
